---
title: "Multiple regression"
author: |
  | Introduction to Quantitative Ecology
  | Fall 2018
  | Chris Sutherland
  | csutherland@umass.edu

header-includes:
   - \usepackage{tabto}
   - \usepackage{verbatim}
   - \usepackage{amsmath}
   - \usepackage{mathtools}
   - \usepackage{graphicx}
   - \usepackage{tikz}
   - \usepackage{tikzpagenodes}
   - \definecolor{OG}{RGB}{0,64,8}
   - \definecolor{LG}{RGB}{0,102,51}
   - \definecolor{myRed}{RGB}{228,26,28}
   - \definecolor{myBlue}{RGB}{55,126,184}
   - \definecolor{myGreen}{RGB}{77,175,74}
   - \definecolor{myPurple}{RGB}{152,78,163}
   - \setbeamercolor{itemize item}{fg=white!0!LG}
   - \setbeamercolor{enumerate item}{fg=white!0!LG}
   - \setbeamercolor{enumerate subitem}{fg=white!70!LG}
   - \setbeamercolor{itemize subitem}{fg=white!70!LG}   
   - \setbeamercolor{itemize subsubitem}{fg=white!70!LG}   
   - \setbeamercolor{navigation symbols}{fg=white!70!LG, bg=white!70!LG}
   - \usepackage{inputenc}
   - \usepackage{booktabs}
   - \usepackage{caption}
   - \usetikzlibrary{patterns,arrows,decorations.pathreplacing}
   - \usepackage{setspace}
   - \DeclareGraphicsExtensions{.pdf,.png,.jpg,.gif}
   - \usepackage{fixmath}
   - \usepackage{tabto}
   - \usepackage{arydshln}
   - \graphicspath{{"G:/Dropbox/C/Courses class/NRC290b/NRC_290b_2018/Images/MultRegres/"}, {"C:/Users/csutherland/Dropbox/C/Courses class/NRC290b/NRC_290b_2018/Images/MultRegres/"}}

output:
  beamer_presentation:
    highlight: tango
    theme: "default"
    colortheme: "spruce"
    fonttheme: "serif"
    slide_level: 2
    incremental: false
classoption: t
---

```{r setup, echo=FALSE, results='hide', message=FALSE}
library(knitr)
opts_knit$set(root.dir = "G:/Dropbox/C/Courses class/NRC290b/NRC_290b_2018/")
#opts_knit$set(root.dir = "C:/Users/csutherland/Dropbox/C/Courses class/NRC290b/NRC_290b_2018/")

options(width=120)
options(width.cutoff=80)
options(scipen = 999)
opts_chunk$set(prompt = FALSE, comment = NA, mysize=TRUE, size="\\scriptsize")
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knit_hooks$set(document = hook2)
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
```

## i-Clicker matching

\vfill

\centering \LARGE \href{https://tinyurl.com/ycl83wa5}{https://tinyurl.com/ycl83wa5}

\vfill

## Regression

What do you remember?

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=0cm] at (current page)
{\includegraphics[height=2in]{QM.png}};
\end{tikzpicture}

```{r,echo=FALSE,results="hide"}
par(oma=c(0,0,0,0),mar=c(4,4,1,1))
set.seed(123)
Explanatory <- round(runif(10,0,10),2)
Response <- round(rnorm(10,10 + 3*Explanatory,5),2)
mod <- lm(Response ~ Explanatory)
clr <- adjustcolor(4,0.25)

pdf("Images/MultRegres/scatter.pdf", height=3,width=3)
plot(Response ~ Explanatory, pch=21, bg=clr, bty="l",las=1, cex=1.5, ylim=c(10,40),  xlim=c(0,10))
dev.off()

pdf("Images/MultRegres/scatterline.pdf", height=3,width=3)
plot(Response ~ Explanatory, pch=21, bg=clr, bty="l",las=1, cex=1.5, ylim=c(10,40),  xlim=c(0,10))
abline(lm(Response~Explanatory),lwd=2,col=4)
dev.off()

pdf("Images/MultRegres/scatterestline.pdf", height=3,width=3)
plot(Response ~ Explanatory, pch=21, bg=clr, bty="l",las=1, cex=1.5, ylim=c(10,40),  xlim=c(0,10))
abline(lm(Response~Explanatory),lwd=2,col=4)
points(x=0,y=mod$coefficient[1],pch=16,cex=1.5,col=2)
arrows(2,mod$coefficient[1] + mod$coefficient[2]*2,
       4,mod$coefficient[1] + mod$coefficient[2]*2,lwd=2,col=2,length=0)
arrows(4,mod$coefficient[1] + mod$coefficient[2]*2,
       4,mod$coefficient[1] + mod$coefficient[2]*4,lwd=2,col=2,length=0)
dev.off()

pdf("Images/MultRegres/hyp.pdf", height=3.5,width=6.5)
par(mfrow=c(1,2))
plot(Response ~ Explanatory, pch=21, bg=clr, bty="l",las=1, cex=1.5, ylim=c(10,40),  xlim=c(0,10), main="Null")
abline(h=mean(Response),lwd=2,col=1)
plot(Response ~ Explanatory, pch=21, bg=clr, bty="l",las=1, cex=1.5, ylim=c(10,40),  xlim=c(0,10), main="Alternative")
abline(h=mean(Response),lwd=1,col=1,lty=2)
abline(lm(Response~Explanatory),lwd=2,col=4)
dev.off()

df <- data.frame(Response=Response, Explanatory=Explanatory)
```

## Regression

What we *should* remember about regression:

- to find the relationship between two continuous variables

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-3cm,yshift=3cm] at (current page.south east)
{\includegraphics[height=2.3in]{scatter.pdf}};
\end{tikzpicture}

## Regression

What we *should* remember about regression:

- to find the relationship between two continuous variables
- estimate the correlation coefficient ($r$)
    - how close are the values to the best fit line

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-3cm,yshift=3cm] at (current page.south east)
{\includegraphics[height=2.3in]{scatterline.pdf}};
\end{tikzpicture}

## Regression

What we *should* remember about regression:

- to find the relationship between two continuous variables
- estimate the correlation coefficient ($r$)
    - how close are the values to the best fit line
    - `r round(cor(Response,Explanatory),2)`

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-3cm,yshift=3cm] at (current page.south east)
{\includegraphics[height=2.3in]{scatterline.pdf}};
\end{tikzpicture}

## Regression

What we *should* remember about regression:

- to find the relationship between two continuous variables
- estimate the correlation coefficient ($r$)
    - how close are the values to the best fit line
- estimate the parameters of the best fit (straight) line
    - $y = a + bX$

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-3cm,yshift=3cm] at (current page.south east)
{\includegraphics[height=2.3in]{scatterline.pdf}};
\end{tikzpicture}

## Regression

What we *should* remember about regression:

- to find the relationship between two continuous variables
- estimate the correlation coefficient ($r$)
    - how close are the values to the best fit line
- estimate the parameters of the best fit (straight) line
    - $y = a + bX$
    - $y$: response variable
    - $a$: intercept
    - $b$: slope
    - $X$: explanatory variable

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-3cm,yshift=3cm] at (current page.south east)
{\includegraphics[height=2.3in]{scatterestline.pdf}};
\end{tikzpicture}

## Regression

What we *should* remember about regression:

- to find the relationship between two continuous variables
- estimate the correlation coefficient ($r$)
    - how close are the values to the best fit line
- estimate the parameters of the best fit (straight) line
    - $y = a + bX$
    - $y$: response variable
    - $a$: `r round(coef(lm(Response~Explanatory))[1],2)`
    - $b$: `r round(coef(lm(Response~Explanatory))[2],2)`
    - $X$: explanatory variable

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-3cm,yshift=3cm] at (current page.south east)
{\includegraphics[height=2.3in]{scatterestline.pdf}};
\end{tikzpicture}

## Regression

Regression analysis provides inference about the slope:

- *null* hypothesis: slope *is not* different from 0
- *alternative* hypothesis: slope *is* different from 0
- how?

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=-2cm] at (current page)
{\includegraphics[height=2in]{hyp.pdf}};
\end{tikzpicture}

## Regression

Regression analysis provides inference about the slope:

- *null* hypothesis: slope *is not* different from 0
- *alternative* hypothesis: slope *is* different from 0
- *p* -value of the slope

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=-2cm] at (current page)
{\includegraphics[height=2in]{hyp.pdf}};
\end{tikzpicture}

## Regression in p`R`actice

- in algebra
    - $y = a + bX$
- in `R` (a linear model)
    - `lm(y ~ X)`

## Regression in p`R`actice

- in algebra
    - $Response = a + b \times Explanatory$
- in `R` (a linear model)
    - `lm(Response ~ Explanatory, data=df)`

```{r}
df
```

## Regression in p`R`actice

```{r,eval=FALSE}
mod <- lm(Response ~ Explanatory, data = df)
summary(mod)
```

## Regression in p`R`actice

```{r}
mod <- lm(Response ~ Explanatory, data = df)
summary(mod)
```

## Regression in p`R`actice

```{r}
mod <- lm(Response ~ Explanatory, data = df)
summary(mod)
```

\begin{tikzpicture}[remember picture,overlay]
\draw[red, thick] (-0.1,2.1) -- (10.9,2.1) -- (10.9,4.4) -- (-0.1,4.4) -- (-0.1,2.1);
\end{tikzpicture}

## A more complex problem

In typical biological studies:

- rarely collect only a single explanatory variable
- interested in *joint* effects
- interested in *interactive* effects
- we can use \textcolor{blue}{\textit{multiple regression}}
    - $>1$ explanatory variable
    - explanatory variables are continuous

## An example - lakes in remote North East Russia

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=-0.3cm] at (current page)
{\includegraphics[height=3in]{Gmapwetlandcomplex.png}};
\end{tikzpicture}

## An example - lakes in remote North East Russia

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=-0.3cm] at (current page)
{\includegraphics[height=3in]{Gmapwetlandcomplexbasic.png}};
\end{tikzpicture}

## An example - lakes in remote North East Russia

What might influence the number of a certain fish species (abundance) in each of these ponds?

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=-1cm] at (current page)
{\includegraphics[height=2in]{QM.png}};
\end{tikzpicture}

## An example - lakes in remote North East Russia

What might influence the number of a certain fish species (abundance) in each of these ponds?

- lake size
- pH
- connectivity
- depth
- human activity (e.g., fishing)
- agricultural run-off
- etc$\ldots$

## An example - lakes in remote North East Russia

I am particularly interested in how:

- the size of the lake (`Area`)
- the amount of pesticides in the water (`Pest`)

influence the the number of fish counted in a lake (`Abund`)

```{r,echo=FALSE}
set.seed(123456)
Pest <- round(runif(10,0,10),2)
Area <- round(runif(10,0,10),2)
Abund <- round(rnorm(10,50 + 3*Area - 2*Pest,3),0)
fish <- data.frame(Abund = Abund,
                   Area = Area,
                   Pest = Pest)
```

```{r}
fish
```

## An example - lakes in remote North East Russia

To visualize relationships for >1 covariate:

- plot response against each variable independently

```{r,echo=FALSE,results="hide"}
pdf("Images/MultRegres/cov.pdf",height=3,width=7)
par(mfrow=c(1,2),oma=c(0,0,0,0),mar=c(4,4,2,2))
plot(Abund ~ Area, main="Abundance ~ Area", pch=21, bg=clr, bty="l",las=1, cex=1.5, data=fish, ylab="Abundance", xlab="Area")
plot(fish$Abund ~ fish$Pest, main="Abundance ~ Pest",pch=21, bg=clr, bty="l",las=1, cex=1.5, data=fish, ylab="Abundance", xlab="Pest")
dev.off()

pdf("Images/MultRegres/cov1.pdf",height=3,width=7)
par(mfrow=c(1,2),oma=c(0,0,0,0),mar=c(4,4,2,2))
plot(fish$Abund ~ fish$Area, main="Abundance ~ Area", pch=21, bg=clr, bty="l",las=1, cex=1.5)
plot(fish$Abund ~ fish$Pest, main="Abundance ~ Pest",pch=21, bg=clr, bty="l",las=1, cex=1.5)
dev.off()
pdf("Images/MultRegres/cov2.pdf",height=3,width=7)
par(mfrow=c(1,2),oma=c(0,0,0,0),mar=c(4,4,2,2))
plot(fish$Abund ~ fish$Area, main="Abundance ~ Area", pch=21, bg=clr, bty="l",las=1, cex=1.5)
plot(fish$Abund ~ fish$Pest, main="Abundance ~ Pest",pch=21, bg=clr, bty="l",las=1, cex=1.5)
dev.off()
```

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=-1cm] at (current page)
{\includegraphics[height=2in]{cov.pdf}};
\end{tikzpicture}

## Simple regression

Test for an effect of `Area` alone:

```{r, eval = F}
mod.area <- lm(Abund ~ Area, data = fish)
summary(mod.area)
```

## Simple regression

Test for an effect of `Area` alone:

```{r}
mod.area <- lm(Abund ~ Area, data = fish)
summary(mod.area)
```

\begin{tikzpicture}[remember picture,overlay]
\draw[red, thick] (-0.1,2.7) -- (10.9,2.7) -- (10.9,3.1) -- (-0.1,3.1) -- (-0.1,2.7);
\draw[red, thick] (-0.1,1.4) -- (10.9,1.4) -- (10.9,1) -- (-0.1,1) -- (-0.1,1.4);
\end{tikzpicture}

## Simple regression

Test for an effect of `Area` alone:

- significant positive effect
    - $\beta_{\text{Area}}=$ `r round(mod.area$coefficient[2],2)`
    - $p=$ `r round(summary(mod.area)$coefficients[2,4],7)`
    - $R^2 =$ `r round(summary(mod.area)$adj.r.squared,2)`
    
## Simple regression

Test for an effect of `Pest` alone:

```{r, eval=F}
mod.pest <- lm(Abund ~ Pest, data = fish)
summary(mod.pest)
```

## Simple regression

Test for an effect of `Pest` alone:

```{r}
mod.pest <- lm(Abund ~ Pest, data = fish)
summary(mod.pest)
```

\begin{tikzpicture}[remember picture,overlay]
\draw[red, thick] (-0.1,2.7) -- (10.9,2.7) -- (10.9,3.1) -- (-0.1,3.1) -- (-0.1,2.7);
\draw[red, thick] (-0.1,1.4) -- (10.9,1.4) -- (10.9,1) -- (-0.1,1) -- (-0.1,1.4);
\end{tikzpicture}

## Simple regression

Test for an effect of `Pest` alone:

- significant negative effect
    - $\beta_{\text{Pest}}=$ `r round(mod.pest$coefficient[2],2)`
    - $p=$ `r round(summary(mod.pest)$coefficients[2,4],7)`
    - $R^2 =$ `r round(summary(mod.pest)$adj.r.squared,2)` 

## Simple regression

- significant positive effect of `Area` 
    - $\beta_{\text{Area}}=$ `r round(mod.area$coefficient[2],2)`
    - $p=$ `r round(summary(mod.area)$coefficients[2,4],7)`
    - $R^2 =$ `r round(summary(mod.area)$adj.r.squared,2)`

- significant negative effect of `Pest`
    - $\beta_{\text{Pest}}=$ `r round(mod.pest$coefficient[2],2)`
    - $p=$ `r round(summary(mod.pest)$coefficients[2,4],7)`
    - $R^2 =$ `r round(summary(mod.pest)$adj.r.squared,2)`

## Multiple regression

*Multiple* regression extends the simple linear model:

- to find the relationship between 1 continuous response variable and two or more explanatory variables

## Multiple regression

*Multiple* regression extends the simple linear model:

- to find the relationship between 1 continuous response variable and two or more explanatory variables
- estimate the overall correlation coefficient ($R^2$)
    - how well the \underline{model} fits the data

## Multiple regression

*Multiple* regression extends the simple linear model:

- to find the relationship between 1 continuous response variable and two or more explanatory variables
- estimate the overall correlation coefficient ($R^2$)
    - how well the \underline{model} fits the data
- estimate the parameters of the best fit relationships
    - $y = a + b_1 X_1 + b_2 X_2 + \ldots$

## Multiple regression

*Multiple* regression extends the simple linear model:

- to find the relationship between 1 continuous response variable and two or more explanatory variables
- estimate the overall correlation coefficient ($R^2$)
    - how well the \underline{model} fits the data
- estimate the parameters of the best fit (straight) line
    - $y = a + b_1 X_1 + b_2 X_2 + \ldots$
    - $y$: response variable
    - $a$: intercept
    - $b_1$ $b_2$: slopes (one for each explanatory variable)
    - $X$'s: explanatory variables

## Multiple regression

*Multiple* regression extends the simple linear model:

- to find the relationship between 1 continuous response variable and two or more explanatory variables
- estimate the overall correlation coefficient ($R^2$)
    - how well the \underline{model} fits the data
- estimate the parameters of the best fit (straight) line
    - $y = a + b_1 X_1 + b_2 X_2 + \ldots$
    - $b_1$ $b_2$: slopes (one for each explanatory variable)
- are the slopes *significant*?
    - Null: no effect of $X_1$ on $y$ (i.e., $b_1 = 0$)
    - Null: no effect of $X_2$ on $y$ (i.e., $b_2 = 0$)
    - test using $p$ -values

## Multiple regression in p`R`actice

- in algebra
    - $y = a + b_1 X_1 + b_2 X_2$
- in `R`
    - `lm(y ~ X1 + X2)`

## Multiple regression in p`R`actice

- in algebra
    - $Abund = a + b_1 \times Area + b_2 \times Pest$
- in `R`
    - `lm(Abund ~ Area + Pest, data=fish)`

```{r}
fish
```

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=1.6cm,yshift=-1.6cm] at (current page)
{\includegraphics[height=1.3in]{cov.pdf}};
\end{tikzpicture}

## Multiple Regression

Conduct a multiple regression for each:

```{r, eval = F}
mod.joint <- lm(Abund ~ Area + Pest, data = fish)
summary(mod.joint)
```

## Multiple Regression

Conduct a multiple regression for each:

```{r}
mod.joint <- lm(Abund ~ Area + Pest, data = fish)
summary(mod.joint)
```

\begin{tikzpicture}[remember picture,overlay]
\draw[red, thick] (-0.1,2.7) -- (10.9,2.7) -- (10.9,3.4) -- (-0.1,3.4) -- (-0.1,2.7);
\draw[red, thick] (-0.1,1.45) -- (10.9,1.45) -- (10.9,1.05) -- (-0.1,1.05) -- (-0.1,1.45);
\end{tikzpicture}

## Regression results

- significant positive effect of `Area` alone (simple regression)
    - $\beta_{\text{Area}}=$ `r round(mod.area$coefficient[2],2)`
    - $p=$ `r round(summary(mod.area)$coefficients[2,4],5)`
    - $R^2 =$ `r round(summary(mod.area)$adj.r.squared,2)`

- significant negative effect of `Pest`  alone (simple regression)
    - $\beta_{\text{Pest}}=$ `r round(mod.pest$coefficient[2],2)`
    - $p=$ `r round(summary(mod.pest)$coefficients[2,4],5)`
    - $R^2 =$ `r round(summary(mod.pest)$adj.r.squared,2)`

## Regression results

- significant positive effect of `Area` alone (simple regression)
    - $\beta_{\text{Area}}=$ `r round(mod.area$coefficient[2],2)`
    - $p=$ `r round(summary(mod.area)$coefficients[2,4],5)`
    - $R^2 =$ `r round(summary(mod.area)$adj.r.squared,2)`

- significant negative effect of `Pest`  alone (simple regression)
    - $\beta_{\text{Pest}}=$ `r round(mod.pest$coefficient[2],2)`
    - $p=$ `r round(summary(mod.pest)$coefficients[2,4],5)`
    - $R^2 =$ `r round(summary(mod.pest)$adj.r.squared,2)`

- significant *joint* effects of `Area` & `Pest` (multiple regression)
    - $\beta_{\text{Area}}=$ `r round(mod.joint$coefficient[2],2)` ($p=$ `r round(summary(mod.joint)$coefficients[2,4],5)`)
    - $\beta_{\text{Pest}}=$ `r round(mod.joint$coefficient[3],2)` ($p=$ `r round(summary(mod.joint)$coefficients[3,4],5)`)
    - $R^2 =$ `r round(summary(mod.joint)$adj.r.squared,2)`


## Group Exercise

\footnotesize

Conduct an analysis in `R` to investigate whether the number of voles captured (`Voles`) is influenced by:

- the % vegetation in each habitat patch, `PercVeg`
- the distance to the nearest road , `Dist2Road`
- data: *vole trapping* 

1. For each predictor variable (i.e., simple linear regression):
    - Produce a figure to visualize the hypothesis you are testing
    - Fit a linear model
    - How each predictor influences the number of voles captured?
    
2. Now fit a multiple regression:
    - Produce a 2-panel figure to visualize the hypotheses
    - Fit a `multiple regression' model
    - How each predictor influences the number of voles captured?
    - Do these results differ to the `univariate' models

3. (*optional*) Use AIC to pick the 'best' model
    - See page 301 in the book
    - Use the `add1()` function

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=4.5cm,yshift=1.8cm] at (current page)
{\includegraphics[height=0.7in]{watervole.jpg}};
\end{tikzpicture}
