---
title: "Week 4: Data Management"
subtitle:  "Session 2"

header-includes:
  \usepackage{multicol}
  
output:
  beamer_presentation:
    
    # highlight: tango
#     theme: "default"
#     colortheme: "spruce"
#     fonttheme: "serif"
#     slide_level: 2
#     incremental: false
# urlcolor: blue
# classoption: t
---

```{r setup, echo=FALSE, results='hide', message=FALSE}
library(knitr)
require(ggplot2)
require(data.table)


iris_dt = data.table(iris)

iris_dt[, sep_wid_jit := jitter(Sepal.Width, factor = 2)]
iris_dt[, sep_len_jit := jitter(Sepal.Length, factor = 2)]
iris_dt[, x_jit := jitter(rep(0, nrow(iris_dt), factor = 2))]
iris_dt

# # opts_knit$set(root.dir = "G:/Dropbox/C/Courses class/NRC290b/NRC_290b_2018/")
# #opts_knit$set(root.dir = "C:/Users/csutherland/Dropbox/C/Courses class/NRC290b/NRC_290b_2018")
# 
# opts_chunk$set(prompt = FALSE, comment = NA, mysize=TRUE, size="\\scriptsize")
# hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
# hook2 <- function(x){ gsub("```\n+```\n", "", x) }
# knit_hooks$set(document = hook2)
# knitr::knit_hooks$set(mysize = function(before, options, envir) {
#   if (before) 
#     return(options$size)
# })
```


# iClicker questions


Which of the following are *not* one of the statistical concepts we have discussed in the last few lectures?

\begin{enumerate}[A]
\item statistics and parameters
\item populations
\item samples
\item variance and standard deviation
\item center and spread
\end{enumerate}




\begin{multicols}{2}
\null \vfill
\vfill \null
\columnbreak
\includegraphics[width = 0.35\textwidth]{../slide_images/iClicker_logo.png}
\end{multicols}



# Announcements

No class on Tuesday!

Chapter 4 has a lot of material.  I'm moving the due date for the pre-class reading assignments to Wednesday of next week (by midnight).



# Distributions

Your Gardener book mentions the Normal distribution in chapter 4, but there is no prior discussion about what *distribution* means!

What is a distribution?

We're going to build an intuitive concept of a distribution from the ground up.




# Distributions

<!-- \null \vfill -->
<!-- \vfill \null -->

<!-- \textit{} -->
\begin{multicols}{2}

\vspace*{\fill}
How variable are the sepals of in a wild population of \textit{Iris setosa}?
\vspace*{\fill}

\columnbreak
\includegraphics[width = 0.45\textwidth]{../slide_images/Mature_flower_diagram.png}
\end{multicols}



```{r, fig.width = 6, fig.height = 2.4, echo = FALSE}
ggplot(iris_dt, aes(x = sep_wid_jit, y = x_jit)) + geom_point() + ylim(-0.04, 0.04) +
  ylab("") + xlab("Sepal width (cm)") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  ggtitle("Iris setosa sepal widths")

```



# Distributions

What is the \textbf{center} of the \textit{Iris setosa} sepal width data?

What is the \textbf{dispersion} or \textbf{spread} of the \textit{Iris setosa} sepal width data?

A *histogram* plot displays the relative frequencies of size *bins*.



```{r, fig.width = 8, fig.height = 4, echo = FALSE}
hist(
  iris_dt$sep_wid_jit,
  freq = TRUE,
  main = "Histogram of Iris setosa sepal width", 
  xlab = "Sepal width (cm)")
```



# Distributions

A **distribution** describes **variability** in a collection of data points.

There are many kinds of distributions.




# The Normal Distribution

You've probably heard of the **Normal** distribution.

It describes a collection of data that are centered around a most likely value, the *mean*.  The likelihood of values away from the center decreases *symmetrically* with the magnitude of the distance from the mean.

It has a familiar bell shape:

```{r, echo = FALSE, fig.asp = 1 / 3}

x = seq(-5, 5, length.out = 1000)
ggplot(data.frame(x = x, y = dnorm(x)), aes(x, y)) + geom_line(size = 2)

```


# Distributions

Different distributions describe different kinds of variability.

We'll encounter many types throughout the course.

Here are 3 of the most famous:



```{r, echo = FALSE, fig.asp = 1/2.5}
x1 = seq(0.01, 2, length.out = 1000)
x2 = 0:10
x3 = seq(0.01, 0.99, length.out = 1000)
h = hist(x2, breaks = x2, plot = FALSE)
h$counts = dpois(x2, 2.3)


par(mfrow = c(1, 3))
plot(x1, dexp(x1), type = "l", main = "Exponential", ylab = "Pr")
plot(x1, dbeta(x3, 9, 5), type = "l", main = "Beta", ylab = "Pr")
plot(h, main = "Poisson", ylab = "probability mass")
```



# Center and Spread

What are some statistics we use to measure center?

How could we describe spread?


# Center and Spread in R

These functions can help:
```{r, eval=FALSE}
mean()
range()
median()
min()
max()
sd()
summary()
```




# Group Activity: Salamanders

We have data for salamanders collected at 4 locations in the Pioneer Valley.

The data include information about:

* Snout to vent length
* Total length
* Sex
* Observation season

You can find the data file and detailed instructions on Moodle.

Today you'll do some exploratory data analysis.  Load the data into R and use some of the functions you know to examine the data.

