---
title: "Week 4: Data Management"
subtitle:  "Session 1"
date: Spring 2020
output: beamer_presentation
header-includes: \usepackage{multicol}
---

```{r setup, echo=FALSE, results='hide', message=FALSE}
require(knitr)
tmp = rprojroot::find_rstudio_root_file()
knitr::opts_chunk$set(root.dir = tmp, error = TRUE)
knitr::opts_knit$set(echo = TRUE, error = TRUE, root.dir = tmp)
source(paste0(tmp, "/data/environment_vars.R"))
rm(tmp)
```


# iClicker quiz: Question 1

Your book mentioned several good reasons to collect information on who, what, where, when with your observations.

Which of the following is NOT a reason to collect all of the Ws?

\begin{enumerate}[A]
\item It means you sampled correctly and without bias
\item It allows the data to be used for multiple purposes
\item It ensures that the data you collect can be checked for accuracy
\item It means that you won’t forget some important aspect of the data
\item It allows someone else to repeat the exercise exactly
\end{enumerate}


\begin{multicols}{2}
\null \vfill
\vfill \null
\columnbreak
\includegraphics[width = 0.35\textwidth]{../slide_images/iClicker_logo.png}

\end{multicols}



# iClicker Quiz: Question 2

Which of the following R concepts has been the most challenging or frustrating for you?

\begin{enumerate}[A]
\item Loading data files into R
\item Strings literals vs. variables
\item Sequences and series
\item Navigating swirl
\end{enumerate}

\begin{multicols}{2}
\null \vfill
\vfill \null
\columnbreak
\includegraphics[width = 0.35\textwidth]{../slide_images/iClicker_logo.png}
\end{multicols}




# Announcements

* I'm experimenting with some alternative ways to build lecture slides.
 * Slides may have different looks and feels over the next several weeks.
 * I will ask for your feedback on the slide styles and content later in the course.

* Notetaker

* Do NOT install the Excel Analysis Pack.  We will not be using Excel for data analysis.  I will provide updated information and instructions for passages in the book the refer to analyses in Excel.


 

# For today

* What did we do last week?
* Data concepts
* Recording data
* Variables
* Group activity time




# Data: some key concepts

Data and *metadata*
"Data without metadata is meaningless."

But.... what is metadata?


When we design data sheets, we should think like R:

##Row format
* Each row is an *observation*.
* Each row contains columns for:
 * Metadata (who, where, etc.)
 * Explanatory variables
 * Response variables


Sampling Units: This is a subtle concept, we'll keep returning to it throughout the course.




# Important concepts in data entry

What are some important questions your data sheets should answer?

* Who, what, where, when
* Variables (there may be multiple variables in each category)
* Notes: unusual conditions, etc.
 
Which of these could be considered *metadata*?

It's better to record all of the data and other observations that could be relevant later, even
if you don't end up using all of the information for your analyses.





# Bears/poops group time

I sincerely apologize for the technical issues last class.  Thank you all for bearing with me.

General questions?

Remember the peer evaluations.  They will be available on Moodle starting Thursday.




# Why do we need to sample?

\begin{center}
\includegraphics[width = 0.88\textwidth]{../slide_images/question_mark_green.png}
\end{center}





# Variables: key concepts

Several pairs of terms are often used to describe different types of variables:

* Independent/dependent
* Predictor/response
* Explanatory/response

What do they mean?

What are some important data types?




# Variables example: Can you identify the predictor and response?

```{r, echo = FALSE, include = FALSE}
dat = read.csv("data/cedar_creek_survey.csv")
require(ggplot2)
gg = ggplot(dat, aes(x = Total.Individuals, y = Distinct.Species.Count))

```
```{r, echo = FALSE, fig.asp = 0.6, warning=FALSE}
# plot(
#   Distinct.Species.Count ~ Total.Individuals, 
#   data = dat,
#   xlab = "Total Number of Individuals",
#   ylab = "Number of Distinct Species",
#   main = "Cedar Creek Rarefaction: Preliminary Data Exploration",
#   pch = 16,
#   cex = 1
#   )


gg +
  geom_point() +
  xlab("Count of Individuals") +
  ylab("Count of Distinct Species") +
  ggtitle("Cedar Creek Rarefaction: Preliminary Data Exploration") +
  theme(text = element_text(size = 22))

```



# What to record?

The metadata:

\begin{enumerate}
\item Who recorded the data
\item What is the *sampling unit*
\item Where was the data recorded
\item When was the data recorded
\end{enumerate}





# What to record?

The data (i.e. the variables):

\begin{enumerate}
  \item Response Variable 
  \begin{enumerate}
    \item the variable of interest you are trying to estimate 
    \item similar to/or sometimes called the “dependent variable”
  \end{enumerate}
  \item Explanatory Variables
  \begin{enumerate}
    \item Variables that influence the response variable
    \item similar to/or sometimes called the “independent variables” or “predictor variable”
  \end{enumerate}
\end{enumerate}




# Some examples of variables

Can you identify the predictor and response variables in these *models*?

From high school math, you might remember the equation for a line:
\begin{equation}
   \label{eq:example-equation}
   y = mx + b
\end{equation}


We'll learn all about statistical models of the form:
\begin{equation}
   \label{eq:example-equation}
   y_{i} = \alpha + \beta_{1} \times x_{i1} + \beta_{2} \times x_{i2} + \epsilon
\end{equation}


Why did I call these equations models?