---
title: "Graphical data exploration"
author: "Michael France Nelson"
date: "Spring, 2020"
output: 
  html_document:
    css: !expr here::here("css", "styles.css")
    number_sections: TRUE
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
require(here)
knitr::opts_chunk$set(echo = TRUE)

rarefaction_data = read.csv(here("data", "rarefaction_data_2020.csv"))

```

# Overview

In this activity, you'll be building you skills with graphical exploration of data in R.

You'll need to use several functions that we haven't talked about in class to complete this activity.

You'll also need some new arguments to modify the look of scatterplots.


# Subsetting data frames

Often you want to use only a *subset* of observations contained in a dataframe.

To accomplish this, you need to decide what criteria you want to use to select those observations you wish to keep.

One way to do this is by using *logical subsetting*.

You've read about logical tests in the swirl tutorials, and now you'll have a chance to practice with some real data.


## Subsetting examples 

In the following examples, I'll use one of R's built-in data sets: `mtcars`.

To load the data into R's memory you need to type:
`data(mtcars)`

To familiarize yourself with the data, use `head()` to view the first six rows:

```{r}
data(mtcars)
head(mtcars)
```


The column called "hp" shows the horsepower of each car in the data set:

```{r}
mtcars$hp
```

How can you retrieve the data for cars that have less than 75 horsepower?

You can use a logical test to show which entries in the `hp` column are less than 75:

```{r}
mtcars$hp < 75
```

How can we use this info to pull out the low horsepower cars?

Use the `subset()` function.  It allows us to keep only the rows of the data frame that match our criteria.

Here's how I would retrieve the low horsepower cars:

```{r}
subset(mtcars, hp < 75)
```

How would you select the cars that have exactly 110 horsepower?

```{r}
subset(mtcars, hp == 110)
```

Note the double equals sign: `==`

The main logical operators in R include:

`==`

`>`

`<`

`>=`

`<=`



# Rarefaction data plots

For this activity, we're going to use the rarefaction data that we collected on the first day of class.

You'll need to download the data file `rarefaction_data_2020.csv` from Moodle and load it into R.

The data should look like this:

```{r, echo=FALSE}
head(rarefaction_data)
```

You'll need to create 5 scatterplots.  All of your scatterplots need to have the number of individuals on the x-axis and the number of distinct species on the y axis.

* **Scatterplot 1**: A plot of all samples
* **Scatterplot 2**: A plot of all samples from site 1
* **Scatterplot 3**: A plot of all samples from site 2
* **Scatterplot 4**: A plot of all samples from site 3
* **Scatterplot 5**: A plot of all samples from site 4

You'll need to use the `subset()` function to retrieve the data rows for individual study sites (sites 1 - 4).  Follow the example with `mtcars` above.


## Scatterplot parameters

When you plot the subsets of the rarefaction data, you might notice that the x-axis doesn't always start at 0.  You'll need to use the `xlim` argument within `plot()` to specify the minimum and maximum values shown in the plot window.

For example, here are two plots of data from site #4.  The first plot uses the default x limits, the second plot specifies that the x-axis should range from 0 to 90

```{r, echo=FALSE}
d4 = subset(rarefaction_data, site == 4)

plot(d4$n_individuals, d4$n_species, xlab = "count of individuals", ylab = "count of species", main = "site 4", pch = 18)

plot(d4$n_individuals, d4$n_species, xlab = "count of individuals", ylab = "count of species", main = "site 4", pch = 18, xlim = c(0, 90))

```
<div class="red">
Note:  The second plot omits two of the data points!
</div>

# Exercises

You'll need to create the 5 scatterplots described above: 1 scatterplot that contains all of the data, and four scatterplots that show the data for the four individual sites.

Each scatterplot must:

* Have appropriate title, x-axis label, and y-axis label.
* The x-axis range must include 0 in each plot.  You need to choose an appropriate upper limit that includes all of the data points.  You'll probably need to try multiple values.
* The y-axis in every plot needs to range from 0 to 32

Each scatterplot is worth 8 points, there are 40 total points available for this exercise.

Your group will paste the 5 scatterplots into a document and upload it to Moodle.










```{r, eval=FALSE, include = FALSE}

head(rarefaction_data)
plot(n_species ~ n_individuals, 
     data = rarefaction_data, 
     col = rarefaction_data$site,
     type = "b", ylim = c(0, 35))


with(subset(rarefaction_data, site == 4),
       {
         plot(n_species ~ n_individuals)
       })





curve(expr = 15 - 6 / x, add = T)
rarefaction_data = unique(rarefaction_data)

names(rarefaction_data)[1] = "site"
# write.csv(rarefaction_data, here("data", "rarefaction_data_2020.csv"))

```



