---
title: "testing"
subtitle:  "Session 1"
date: "`r format (Sys.time(), format='%Y-%m-%d %H:%M:%S %z')`"
output:
 beamer_presentation:
  pandoc_args: !expr paste0(here::here('css', 'beamer.yaml'))
header-includes: 
 \input{`r here::here("css", "headers_tikz.yaml")`}
---


\newcommand\iClick{
\vfill
\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2cm,yshift=1cm] at (current page.south east)
{\includegraphics[height=0.25in]{../slides/slide_images/iClicker_logo.png}};
\end{tikzpicture}
}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(rmarkdown)
```


## as
\iClick

## iClicker logo

```{r cars}
summary(cars)
```

\vfill
\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2cm,yshift=1cm] at (current page.south east)
{\includegraphics[height=0.25in]{../slides/slide_images/iClicker_logo.png}};
\end{tikzpicture}


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
