---
title: "testing"
subtitle:  "Session 1"
date: "`r format (Sys.time(), format='%Y-%m-%d %H:%M:%S %z')`"
output:
  beamer_presentation:
    pandoc_args: !expr paste0(here::here('css', 'beamer.yaml'))
header-includes: 
  \input{`r here::here("css", "headers_tikz.yaml")`}
---

<!-- \newcommand\{\iclick}{% -->
<!-- \vfill -->
<!-- \begin{tikzpicture}[remember picture,overlay] -->
<!-- \node[xshift=-2cm,yshift=1cm] at (current page.south east) -->
<!-- {\includegraphics[height=0.25in]{`r here::here("slides", "slide_images", "iClicker_logo.png")`"}}; -->
<!-- \end{tikzpicture}} -->

<!-- \newcommand\{iClick}{ -->
<!-- \vfill -->
<!-- \begin{tikzpicture}[remember picture,overlay] -->
<!-- \node[xshift=-2cm,yshift=1cm] at (current page.south east) -->
<!-- {\includegraphics[height=0.25in]{../slides/slide_images/iClicker_logo.png}}; -->
<!-- \end{tikzpicture} -->
<!-- } -->


# newest test slide
<!-- \vfill -->
\newcommand{\atest}{
 \begin{tikzpicture}[remember picture,overlay]
  \node[xshift=-2cm,yshift=1cm] at (current page.south east){
   \includegraphics[height=0.25in]{`r here::here("slides", "slide_images", "iClicker_logo.png")`"}
  };
\end{tikzpicture}
}

\atest

# newest test slide


<!-- https://tex.stackexchange.com/questions/32661/how-to-locate-figures-with-x-y-specified-location-in-a-presentation/330626 -->

<!-- \begin{tikzpicture}[remember picture,overlay] -->
<!--  \node[anchor=south west,inner sep=0pt] at ((current page.south west) + (2cm,5cm)) { -->
<!--    \includegraphics[height=0.25in]{`r here::here("slides", "slide_images", "iClicker_logo.png")`} -->
<!--  }; -->
<!-- \end{tikzpicture} -->



# Chris's way

\vfill
\begin{tikzpicture}[remember picture,overlay]
 \node[xshift=-2cm,yshift=1cm] at (current page.south east){
  \includegraphics[height=0.25in]{`r here::here("slides", "slide_images", "iClicker_logo.png")`"}
 };
\end{tikzpicture}


# newest test slide

\begin{tikzpicture}
\includegraphics[height=0.25in]{`r here::here("slides", "slide_images", "iClicker_logo.png")`}
\end{tikzpicture}



# break


\newcommand{\latest}
{
 and seven years ago
}

abc \latest


\newcommand{\atest}
{
\includegraphics[height=0.25in]{../slides/slide_images/iClicker_logo.png}
}



\atest

<!-- ## new slide -->

<!-- \includegraphics[height=0.25in]{`r here::here("slides", "slide_images", "iClicker_logo.png")`} -->


<!-- \newcommand{\atest} -->
<!-- { -->
<!-- \includegraphics[height=0.25in]{`r here::here("slides", "slide_images", "iClicker_logo.png")`} -->
<!-- } -->
<!-- \atest -->







<!-- \newcommand{\atest} -->
<!-- { -->
<!-- \includegraphics[height=0.25in]{`r here::here("slides", "slide_images", "iClicker_logo.png")`} -->
<!-- \includegraphics[height=0.25in]{`r here::here("slides", "slide_images", "iClicker_logo.png")`} -->
<!-- } -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# require(rmarkdown)
```


## as
<!-- \iClick -->

## iClicker logo

```{r cars}
# summary(cars)
```

<!-- \vfill -->
<!-- \begin{tikzpicture}[remember picture,overlay] -->
<!-- \node[xshift=-2cm,yshift=1cm] at (current page.south east) -->
<!-- {\includegraphics[height=0.25in]{`r here::here("slides", "slide_images", "iClicker_logo.png")`"}}; -->
<!-- \end{tikzpicture} -->



<!-- {\includegraphics[height=0.25in]{../slides/slide_images/iClicker_logo.png}}; -->

