---
title: "Week 6: Graphical Data Exploriation"
subtitle:  "Session 2"
date: Spring 2020
output:
  beamer_presentation:
    pandoc_args: !expr paste0(here::here('css', 'beamer.yaml'))
header-includes:
  \input{`r here::here("css", "headers_tikz.tex")`}
---

```{r setup, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
require(knitr)
require(here)
source(here("data", "environment_vars.R"))

i = 1;
set.seed(12345)
nrow = 10
dat = data.frame(length = runif(n = nrow), width = rexp(nrow), mass = rpois(nrow, 34))
```




## iClicker Question `r colorize(i, "blue")` `r i = i + 1`

Which of the following should I use to read the file `mydata.csv` into a data frame called `dat` in R?

\begin{enumerate}[A]
\item \texttt{dat = read.csv(mydata.csv)}
\item \texttt{read.csv(mydata.csv)}
\item \texttt{dat = read.csv("mydata.csv")}
\item \texttt{read.csv(mydata.csv, row.names = 1)}
\item \texttt{read.csv("mydata.csv")}
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}



## iClicker Question `r colorize(i, "blue")` `r i = i + 1`

Which of the following lines of code will make a scatterplot with a custom **title**?

```{r, echo = FALSE}
head(dat, 3)
```

\begin{enumerate}[A]
\item \texttt{plot(dat\$mass, dat\$length, col = 3)}
\item \texttt{plot(dat\$mass, dat\$length, type = "l")}
\item \texttt{plot(dat\$length, dat\$mass, type = "p")}
\item \texttt{plot(dat\$mass, dat\$length, main = "data plot")}
\item \texttt{plot(dat\$length, dat\$mass)}
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}



## iClicker Question `r colorize(i, "blue")` `r i = i + 1`

Which is the best location for me to type my R code for in-class and individual activities?

\begin{enumerate}[A]
\item A text file in Word or another word processor
\item The RStudio console pane
\item The RStudio code editor pane
\item The RStudio Environment pane
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}




## Announcements

- Error in Tuesday's title slide is now fixed.  You can re-download the correct version on Moodle.




## Follow-up questions from the Chapter 5 homework

Question 1: What is the difference between parametric and a non-parametric distribution?

- The book's explanation of *parametric distribution* is slightly misleading.  The reading characterizes the Normal distribution as a parametric distribution, however there are many *parametric distributions*. 
- I prefer to use the term *theoretical distributions* when I'm describing *parametric distributions* to emphasize that they are precisely mathematically defined.
- The behavior of *parametric distributions* is governed by 1 or more *parameters* in the probability functions.

- What is a **non-parametric distribution**?




## Follow-up questions from the Chapter 5 homework

Question 3: In R, what would be returned if you compute the mean value of a vector containing the following values: 4, 9, 2, 13, NA, 9? Please describe your methods.

- Question from me:  Is there a benefit to having `mean()` fail when you pass an NA?




## For Today

- Are there any general questions about data exploration, numeric or graphical?
- Short lecture
- Group activity: graphical data exploration: rarefaction data.
- We'll start to use *inference* in the coming weeks!



## Some graphics pointers

In summary, graphs are a useful data visualization tool

- summarizing
- understanding
- describing
- presenting/communicating




## Some graphics pointers

In summary, graphs are a useful data visualization tool

- summarizing
- understanding
- describing
- presenting/communicating

**BUT** we must label them well or they are useless.

- label both axes
- provide a main title for your graph
- avoid clutter
- make it readable
- *I expect graphs to be properly labeled from now on*!




## Some graphics pointers

In summary, graphs are a useful data visualization tool

\vspace{1cm}

|Purpose | Graph Type|
|--------|-----------|
| Illustrating *distribution* | Histogram, Density plot |
|                             | Box(-whisker) plot      |
| Illustrating *differences*  | Bar chart, Box plot     |
| Illustrating *correlations* | Scatter plot            |
| Illustrating *associations* | Pie chart, Bar chart    |
| Illustrating *sample size*  | Line plot of running avg|




## Beyond graphs, Towards statistics

- Graphs are powerful tools that provide insight and understanding of the patterns and relationships in the data.

- Graphs alone don't give us the complete answer.  We need to **quantify** the relationships we see in our plots.




## Beyond graphs, Towards statistics

- How can we **quantify** our evidence for relationships?
    - Are differences between groups *significant*?
    - Are differences between groups *meaningful*?
    
\begin{tikzpicture}[remember picture,overlay]
\node[yshift = 4cm] at (current page.south)
{\includegraphics[height = 7cm]{`r here("slides", "slide_images", "salbox.pdf")`}};
\end{tikzpicture}




## Beyond graphs, Towards statistics

- How can we **quantify** our evidence for relationships?
    - Are associations between two variables *significant*?
    - Are associations between two variables *meaningful*?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift = 4cm] at (current page.south)
{\includegraphics[height = 7cm]{`r here("slides", "slide_images", "salscatter.pdf")`}};
\end{tikzpicture}




## Beyond graphs, Towards statistics

- Graphs are powerful tools that provide insight and understanding of the patterns and relationships in the data.

- Graphs alone don't give us the complete answer.  We need to **quantify** the relationships we see in our plots.

- Statistics is the tool we use to formally answer these questions:
    - Are differences *significant*?
    - Are associations *significant*?
    



## Group graphical activity

- Plotting the rarefaction data
- See the instructions on Moodle

