---
title: "Week 6: Graphical Data Exploration"
subtitle:  "Session 1"
date: Spring 2020
output:
  beamer_presentation:
    pandoc_args: !expr paste0(here::here('css', 'beamer.yaml'))
header-includes:
  \input{`r here::here("css", "headers_tikz.tex")`}
---


```{r setup, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
require(knitr)
require(here)
source(here("data", "environment_vars.R"))
knitr::opts_chunk$set(root.dir = here(), error = TRUE)
knitr::opts_knit$set(echo = TRUE, error = TRUE, root.dir = here())
i = 1
set.seed(12345)
nrow = 10
dat = data.frame(length = runif(n = nrow), width = rexp(nrow), mass = rpois(nrow, 34))
```




## iClicker Question `r colorize(i, "blue")` `r i = i + 1`

Which of the following should I use to read the file `mydata.csv` into a data frame called `dat` in R?

\begin{enumerate}[A]
\item \texttt{dat = read.csv(mydata.csv)}
\item \texttt{read.csv(mydata.csv)}
\item \texttt{dat = read.csv("mydata.csv")}
\item \texttt{read.csv(mydata.csv, row.names = 1)}
\item \texttt{read.csv("mydata.csv")}
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}




## iClicker Question `r colorize(i, "blue")` `r i = i + 1`

Which symbol do we use to represent the **sample mean**?

\begin{enumerate}[A]
\item $\sigma$
\item $\bar s$
\item $\bar{x}$
\item $\mu$
\item $\bar{m}$
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}




## iClicker Question  `r colorize(i, "blue")` `r i = i + 1`

Which symbol do we use to represent the **population mean**?

\begin{enumerate}[A]
\item $\sigma$
\item $\bar s$
\item $\bar{x}$
\item $\mu$
\item $\bar{m}$
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}




## iClicker Question `r colorize(i, "blue")` `r i = i + 1`

Which plot type is most appropriate to show the **distribution** of a set of measurements?

\begin{enumerate}[A]
\item scatterplot
\item boxplot
\item barchart
\item histogram
\item pie chart
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}




## iClicker Question `r colorize(i, "blue")` `r i = i + 1`

Which of the following lines of code will make a scatterplot of the dataframe with length on the x-axis and mass on the y-axis?

```{r, echo = FALSE}
head(dat, 3)
```

\begin{enumerate}[A]
\item \texttt{plot(dat\$mass, dat\$length)}
\item \texttt{scatter(dat\$length, dat\$mass)}
\item \texttt{boxplot(dat\$length, dat\$mass, type = "p")}
\item \texttt{dotplot(dat\$length, dat\$mass)}
\item \texttt{plot(dat\$length, dat\$mass)}
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}




## Announcements

Trying a different slide layout today




## Graphical exploration

Why use graphs?

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=0cm] at (current page)
{\includegraphics[height=2in]{`r here(slide_img_dir, "question_mark_green.png")`}};
\end{tikzpicture}




## Graphical exploration

Two main reasons to use graphs:

\begin{itemize}
\item[1.]<1-> Inform how to analyze the data
  \begin{itemize}
  \item<2-> visualization
  \item<2-> identify patterns
  \item<2-> choose appropriate statistical test
  \end{itemize}
\item[2.]<1-> Presentation of the data
  \begin{itemize}
  \item<3-> summarize results
  \item<3-> communicate results
  \item<3-> publish results
  \end{itemize}
\end{itemize}

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=-2.8cm] at (current page)
{\includegraphics[height=1in]{`r here(chris_img_dir, "graffs.png")`}};
\end{tikzpicture}




## Types of graphs - \emph{Exploratory}

Exploratory graphs help understand the distribution of the data:

- are the data normally distributed?
    - Normality is an important assumption in statistics
    - Normality determines how data are analyzed
- what is the central tendency?
- what is the spread?
- general summaries of the data




## Exploratory: \emph{Histogram}

- width of bars are defined data bins or intervals
- height of bars represent bin-specific frequencies

```{r, eval=FALSE}
hist(values)
```
\begin{tikzpicture}[remember picture,overlay]
\node[yshift=2cm] at (current page.south)
{\includegraphics[height=1.9in]{`r here(slide_img_dir, "randhist.pdf")`}};
\end{tikzpicture}




## Exploratory: \emph{Histogram}

- width of bars are defined data bins or intervals
- height of bars represent bin-specific frequencies

You can change the number and widths of the bins.

```{r, eval = FALSE}
hist(values)
hist(values, breaks = seq(from = 10, to = 90, by = 2))
```
\begin{tikzpicture}[remember picture,overlay]
\node[yshift=2cm] at (current page.south)
{\includegraphics[height=1.9in]{`r here(slide_img_dir, "randhist2.pdf")`}};
\end{tikzpicture}




## Exploratory: \emph{Histogram + Density Plot}

A *density plot*: smoothed version of histogram

- To overlay on a histogram, tell `hist()` to plot the *probability* version of the histogram:

```{r, eval = FALSE}
hist(values, probability = TRUE)
lines(density(values))
```

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=2cm] at (current page.south)
{\includegraphics[height=1.9in]{`r here(slide_img_dir, "randhist3.pdf")`}};
\end{tikzpicture}




## Exploratory: \emph{Box-whisker/Box plot}

- distribution
- outliers
- symmetry or skewness

```{r, eval = FALSE}
boxplot(values)
```

\begin{tikzpicture}[overlay]
\node[inner sep=0pt] (one) at (5,-2.2) {\includegraphics[scale=0.75]{`r here(slide_img_dir, "bw.pdf")`}};
\end{tikzpicture}




## Exploratory: \emph{Box-whisker/Box plot}

- `R`: \texttt{boxplot(x)}  \texttt{\textcolor{gray}{\# x is data}}

\begin{tikzpicture}[overlay]
\node[inner sep=0pt] (one) at (5,-2.7) {\includegraphics{`r here(slide_img_dir, "four.pdf")`}};
\draw[thick,blue,<-] (6.5,-2.7) -- (7.5,-2.7) node[right, black] {Median ($Q_{2}$)};
\draw[thick,blue,<-] (6.5,-2.1) -- (7.5,-2.1) node[right, black] {Upper quartile ($Q_{3}$)};
\draw[thick,blue,<-] (6.5,-3.25) -- (7.5,-3.25) node[right, black] {Lower quartile ($Q_{1}$)};
\draw[thick,blue,<->] (4.5,-3.25) -- (4.5,-2.1) node[midway, left, black, text width = 2.5cm] {Inter quartile range (IQR)};
\draw[thick,blue,<-] (5,-0.38) -- (4,-0.38) node[left, black] {Maximum};
\draw[thick,blue,<-] (5,-4.9) -- (4,-4.9) node[left, black] {Minimum};
\end{tikzpicture}




## Exploratory: \emph{Line graph}

Line graph is a useful plot for running average or time series data

```{r echo=T, eval=F}
# "l": line, "p": points, "b": both
plot(bear.run, type = "l") 
lines(poop.run)
```

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=3cm] at (current page.south) {\includegraphics[height=2.5in]{`r here(slide_img_dir, "runavg.pdf")`}};
\end{tikzpicture}




## Differences

To visualize differences between groups

- box-whisker plots
    - compares averages
    - compares distribution
- bar charts
    - compares averages




## Differences: \emph{Box-whisker plot}

Compare salamander snout-vent lengths by three sexes:

```{r eval=F}
boxplot(mander$SVL ~ mander$Sex) #formula notation
```

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=4cm] at (current page.south)
{\includegraphics[height=3in]{`r here(slide_img_dir, "salbox.pdf")`}};
\end{tikzpicture}




## Differences: \emph{Bar chart}

Compare salamander snout-vent lengths by three sexes:

```{r eval=F}
bars <- tapply(mander$SVL,mander$Sex,mean) #create matrix
barplot(bars)                              # plot it
```

\begin{tikzpicture}[remember picture,overlay]
\node[yshift = 3.1cm] at (current page.south)
{\includegraphics[height=3in]{`r here(slide_img_dir, "salbar.pdf")`}};
\end{tikzpicture}




## Differences: \emph{Bar chart} with associated error

Compare salamander snout-vent lengths by three sexes:

```{r eval=F}
bars <- tapply(mander$SVL,mander$Sex,mean)
barplot(bars)
```

\begin{tikzpicture}[remember picture,overlay]
\node[yshift = 3.1cm] at (current page.south)
{\includegraphics[height=3in]{`r here(slide_img_dir, "salbarse.pdf")`}};
\end{tikzpicture}




## Links

Two main approaches for relationships between data:

\begin{itemize}
\item[1.]<1-> Correlations
\item[2.]<1-> Associations
\end{itemize}




## Links

Two main approaches for graphing relationships between data:

\vspace{0.06cm}

\begin{itemize}
\item[1.]<1-> Correlations
  \begin{itemize}
  \item<1-> two numeric variables
    \begin{itemize}
    \item<1-> \emph{de}pendent variable (of primary interest: y-axis)
    \item<1-> \emph{inde}pendent variable (explanatory variable: x-axis)
    \end{itemize}
  \item<1-> how one variable is related to another
  \item<1-> \emph{scatter plots}
  \end{itemize}
\end{itemize}




## Links: \emph{Scatter plot}

```{r eval=F}
plot(x,y) # x and y are numeric vectors
```

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=4cm] at (current page.south)
{\includegraphics[height=2.5in]{`r here(slide_img_dir, "salscatter.pdf")`}};
\end{tikzpicture}




## Links

Two main approaches for graphing relationships between data:

\begin{itemize}
\item[2.]<1-> Associations
  \begin{itemize}
  \item<1-> categorical data
  \item<1-> summarize categories
    \begin{itemize}
    \item<1-> counts
    \item<1-> proportions
    \item<1-> by rows and/or columns of a table
  \end{itemize}
  \item<1-> \emph{pie charts} for single categories
  \item<1-> \emph{bar graphs} for several categories
  \end{itemize}
\end{itemize}




## Links: \emph{Pie chart}

```{r eval=F}
pietab <- table(classData$Eyes)
pie(pietab) # number of people with each eye color
```

\begin{tikzpicture}[remember picture,overlay]
\node[yshift = 3.5cm] at (current page.south)
{\includegraphics[height=2in]{`r here(slide_img_dir, "classpie.pdf")`}};
\end{tikzpicture}




## Links: \emph{Bar chart}

```{r eval=F}
bartab <- table(classData$Gender, classData$Eyes)
#(number of each gender with each eye color)
barplot(pietab, beside=TRUE)
```

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=3.5cm] at (current page.south)
{\includegraphics[height=2.5in]{`r here(slide_img_dir, "classbar.pdf")`}};
\end{tikzpicture}




## Some graphics pointers

In summary, graphs are a useful data visualization tool

- summarizing
- understanding
- describing
- presenting/communicating




## Some graphics pointers

In summary, graphs are a useful data visualization tool

- summarizing
- understanding
- describing
- presenting/communicating

**BUT** we must label them well or they are useless!

- label both axes
- provide a main title for your graph
- avoid clutter
- make it readable
- *I expect graphs to be properly labeled from now on*!




## Some graphics pointers

In summary, graphs are a useful data visualization tool

\vspace{1cm}

|Purpose | Graph Type|
|--------|-----------|
| Illustrating *distribution* | Histogram, Density plot |
|                             | Box(-whisker) plot      |
| Illustrating *differences*  | Bar chart, Box plot     |
| Illustrating *correlations* | Scatter plot            |
| Illustrating *associations* | Pie chart, Bar chart    |
| Illustrating *sample size*  | Line plot of running avg|



## Beyond graphs, Towards statistics

- Graphs are powerful tools that provide insight and understanding of the patterns and relationships in the data.

- Graphs alone don't give us the complete answer.  We need to **quantify** the relationships we see in our plots.




## Beyond graphs, Towards statistics

- How can we **quantify** our evidence for relationships?
    - Are differences between groups *significant*?
    - Are differences between groups *meaningful*?
    
\begin{tikzpicture}[remember picture,overlay]
\node[yshift = 4cm] at (current page.south)
{\includegraphics[height = 7cm]{`r here("slides", "slide_images", "salbox.pdf")`}};
\end{tikzpicture}




## Beyond graphs, Towards statistics

- How can we **quantify** our evidence for relationships?
    - Are associations between two variables *significant*?
    - Are associations between two variables *meaningful*?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift = 4cm] at (current page.south)
{\includegraphics[height = 7cm]{`r here("slides", "slide_images", "salscatter.pdf")`}};
\end{tikzpicture}




## Beyond graphs, Towards statistics

- Graphs are powerful tools that provide insight and understanding of the patterns and relationships in the data.

- Graphs alone don't give us the complete answer.  We need to **quantify** the relationships we see in our plots.

- Statistics is the tool we use to formally answer these questions:
    - Are differences *significant*?
    - Are associations *significant*?
    

