---
title: "Correlations"
author: |
  | Introduction to Quantitative Ecology
  | Fall 2018
  | Chris Sutherland
  | csutherland@umass.edu

header-includes:
   - \usepackage{tabto}
   - \usepackage{verbatim}
   - \usepackage{amsmath}
   - \usepackage{mathtools}
   - \usepackage{graphicx}
   - \usepackage{tikz}
   - \usepackage{tikzpagenodes}
   - \definecolor{OG}{RGB}{0,64,8}
   - \definecolor{LG}{RGB}{0,102,51}
   - \definecolor{myRed}{RGB}{228,26,28}
   - \definecolor{myBlue}{RGB}{55,126,184}
   - \definecolor{myGreen}{RGB}{77,175,74}
   - \definecolor{myPurple}{RGB}{152,78,163}
   - \setbeamercolor{itemize item}{fg=white!0!LG}
   - \setbeamercolor{enumerate item}{fg=white!0!LG}
   - \setbeamercolor{enumerate subitem}{fg=white!70!LG}
   - \setbeamercolor{itemize subitem}{fg=white!70!LG}   
   - \setbeamercolor{itemize subsubitem}{fg=white!70!LG}   
   - \setbeamercolor{navigation symbols}{fg=white!70!LG, bg=white!70!LG}
   - \usepackage{inputenc}
   - \usepackage{booktabs}
   - \usepackage{caption}
   - \usetikzlibrary{patterns,arrows,decorations.pathreplacing}
   - \usepackage{setspace}
   - \DeclareGraphicsExtensions{.pdf,.png,.jpg,.gif}
   - \usepackage{fixmath}
   - \usepackage{tabto}
   - \usepackage{arydshln}
   - \graphicspath{{"G:/Dropbox/C/Courses class/NRC290b/NRC_290b_2018/Images/Corr/"}, {"C:/Users/csutherland/Dropbox/C/Courses class/NRC290b/NRC_290b_2018/Images/Corr/"}}

output:
  beamer_presentation:
    highlight: tango
    theme: "default"
    colortheme: "spruce"
    fonttheme: "serif"
    slide_level: 2
    incremental: false
classoption: t
---

```{r setup, echo=FALSE, results='hide', message=FALSE}
library(knitr)
opts_knit$set(root.dir = "G:/Dropbox/C/Courses class/NRC290b/NRC_290b_2018/")
#opts_knit$set(root.dir = "C:/Users/csutherland/Dropbox/C/Courses class/NRC290b/NRC_290b_2018/")

options(width=120)
options(width.cutoff=80)
opts_chunk$set(prompt = FALSE, comment = NA, mysize=TRUE, size="\\scriptsize")
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knit_hooks$set(document = hook2)
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
```

## Group evaluations

1. We would *not* use the Spearman's rank test to calculate a correlation coefficient - which one?

```{r,echo=F, results='hide'}
pdf("Images//Corr//corQuiz1.pdf",height=3,width=8.5)
par(mfrow=c(1,3),mar=c(2,2,4,1),oma=c(0,0,0,0))
set.seed(3)
plot(log(1:20),rlnorm(20,seq(0,0.7,length=20)*2,0.1),main="A",axes=F,xlab="",ylab="",pch=21,bg=adjustcolor(4,0.4),cex=2,cex.main=2)
box(bty="l")
plot(1:20,rnorm(20,seq(0,0.7,length=20)*2,0.1),main="B",axes=F,xlab="",ylab="",pch=21,bg=adjustcolor(4,0.4),cex=2,cex.main=2)
box(bty="l")
plot(1:20,rnorm(20,seq(0,0.7,length=20)*3 - seq(0,0.7,length=20)^2*5,0.1),main="C",axes=F,xlab="",ylab="",pch=21,bg=adjustcolor(4,0.4),cex=2,cex.main=2)
box(bty="l")
dev.off()
```


\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=-1cm] at (current page)
{\includegraphics[height=1.5in]{corQuiz1.pdf}};
\end{tikzpicture}

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2cm,yshift=1cm] at (current page.south east)
{\includegraphics[height=0.25in]{iclicker.png}};
\end{tikzpicture}


## Group evaluations

2. For her PhD, Eugene is studying the effects of annual temperature on American robin reproductive success (number of eggs hatched). What is the *dependent* variable? 

\begin{enumerate}[\hspace{0.5cm}A.]
  \item[A)] American robin
  \item[B)] Eugene
  \item[C)] Temperature
  \item[D)] Number of eggs
\end{enumerate}

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2cm,yshift=1cm] at (current page.south east)
{\includegraphics[height=0.25in]{iclicker.png}};
\end{tikzpicture}

## Group evaluations

3. What is the most likely Pearson's correlation coefficient ($r$) for this relationship?

\begin{enumerate}[\hspace{0.5cm}A.]
  \item[A)] 0.90
  \item[B)] 0.09
  \item[C)] -0.9
  \item[D)] -0.09
\end{enumerate}

\vfill
\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-4cm, yshift=-1cm] at (current page.east)
{\includegraphics[height=2in]{corrr.pdf}};
\end{tikzpicture}

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2cm,yshift=1cm] at (current page.south east)
{\includegraphics[height=0.25in]{iclicker.png}};
\end{tikzpicture}

## Group evaluations

4. What is the slope in this equation of a straight line?

$$y = mx + c$$

\begin{enumerate}[\hspace{0.5cm}A.]
  \item[A)] $y$
  \item[B)] $m$
  \item[C)] $x$
  \item[D)] $c$
\end{enumerate}

\vfill

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2cm,yshift=1cm] at (current page.south east)
{\includegraphics[height=0.25in]{iclicker.png}};
\end{tikzpicture}

## Group evaluations

\vfill

5. Which of the following is *polynomial* relationship?

\begin{enumerate}[\hspace{0.5cm}A.]
  \item[A)] $y = ax + c$
  \item[B)] $y = ax + bx^2 + c$
  \item[C)] $y = log(ax) + c$
  \item[D)] $y = c$
\end{enumerate}

\vfill

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2cm,yshift=1cm] at (current page.south east)
{\includegraphics[height=0.25in]{iclicker.png}};
\end{tikzpicture}

```{r,eval=TRUE,echo=FALSE,results="hide"}
x1 <- rnorm(100000,50,10)
x2 <- rnorm(100000,150,10)
x3 <- rnorm(100000,52,10)
x4 <- rnorm(100000,60,10)

pdf("Images//Corr//meansA.pdf",height=4,width=5)
par(mfrow=c(2,1),oma=c(0,0,0,0),mar=c(2,2,1,1))
plot(density(x1),col=1,lwd=1,xlim=c(0,200),ylim=c(0,0.05),
     main="",axes=F,xlab="",ylab="")
abline(v=mean(x1),lwd=2,col=2)
abline(v=mean(x2),lwd=2,col=4)
lines(density(x1),col=1, lwd=1)
lines(density(x2),col=1, lwd=1)
axis(1)
box(bty="l")
boxplot(cbind(x1,x2),ylim=c(0,200),horizontal=TRUE,pch="",
        col=adjustcolor(c(2,4),0.3),axes=F)
axis(1)
abline(v=mean(x1),lwd=2,col=2)
abline(v=mean(x2),lwd=2,col=4)
boxplot(cbind(x1,x2),ylim=c(0,200),horizontal=TRUE,pch="",
        col=adjustcolor(c(2,4),0.3),axes=F,add=T)
box(bty="l")
dev.off()

pdf("Images//Corr//means2.pdf",height=4,width=5)
par(mfrow=c(2,1),oma=c(0,0,0,0),mar=c(2,2,1,1))
plot(density(x1),col=1,lwd=1,xlim=c(0,200),ylim=c(0,0.05),
     main="",axes=F,xlab="",ylab="")
abline(v=mean(x1),lwd=2,col=2)
abline(v=mean(x3),lwd=2,col=4)
lines(density(x1),col=1, lwd=1)
lines(density(x3),col=1, lwd=1)
axis(1)
box(bty="l")
boxplot(cbind(x1,x3),ylim=c(0,200),horizontal=TRUE,pch="",
        col=adjustcolor(c(2,4),0.3),axes=F)
axis(1)
abline(v=mean(x1),lwd=2,col=2)
abline(v=mean(x3),lwd=2,col=4)
boxplot(cbind(x1,x3),ylim=c(0,200),horizontal=TRUE,pch="",
        col=adjustcolor(c(2,4),0.3),axes=F,add=T)
box(bty="l")
dev.off()

pdf("Images//Corr//means3.pdf",height=4,width=5)
par(mfrow=c(2,1),oma=c(0,0,0,0),mar=c(2,2,1,1))
plot(density(x1),col=1,lwd=1,xlim=c(0,200),ylim=c(0,0.05),
     main="",axes=F,xlab="",ylab="")
abline(v=mean(x1),lwd=2,col=2)
abline(v=mean(x4),lwd=2,col=4)
lines(density(x1),col=1, lwd=1)
lines(density(x4),col=1, lwd=1)
axis(1)
box(bty="l")
boxplot(cbind(x1,x4),ylim=c(0,200),horizontal=TRUE,pch="",
        col=adjustcolor(c(2,4),0.3),axes=F)
axis(1)
abline(v=mean(x1),lwd=2,col=2)
abline(v=mean(x4),lwd=2,col=4)
boxplot(cbind(x1,x4),ylim=c(0,200),horizontal=TRUE,pch="",
        col=adjustcolor(c(2,4),0.3),axes=F,add=T)
box(bty="l")
dev.off()
```

## Correlations

- What are examples of correlations? 
- Why would we be interested in correlations?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1cm] at (current page)
{\includegraphics[height=2in]{QM.png}};
\end{tikzpicture}

## Correlations

Interested in the relationship between two samples

- Dependent variable:
    - data we are interested in explaining
    - Y-axis
    - *response* variable
    
```{r,eval=TRUE,echo=FALSE,results="hide"}
set.seed(12)
AR <- runif(25,0,10)
SR <- rnorm(25,10 + 2*AR,2)
pdf("Images//Corr//specareaA.pdf",height=3,width=4)
par(mfrow=c(1,1),oma=c(0,0,0,0),mar=c(4,4,1,1),mgp=c(1,1,1))
plot(AR,SR,type="p",bg=adjustcolor(4,0.4),xlab="",ylab="Dependent variable",ylim=c(5,30),las=1, bty="l",pch=21, cex=2, axes=F)
box(bty="l")
dev.off()

set.seed(1)
x1 <- runif(25,0,10)
y1 <- rnorm(25,10 + 5*x1,6)
pdf("Images//Corr//specareaWith.pdf",height=3,width=4)
par(mfrow=c(1,1),oma=c(0,0,0,0),mar=c(4,4,1,1))
plot(x1,y1,type="p",bg=adjustcolor(4,0.4),ylab="Species Richness",xlab="Area (ha)",las=1, bty="l",pch=21, cex=2, axes=F, ylim=c(0,70))
box(bty="l")
axis(2, las=1)
axis(1,at=1:10)
dev.off()

pdf("Images//Corr//specareaWithout.pdf",height=3,width=4)
par(mfrow=c(1,1),oma=c(0,0,0,0),mar=c(4,4,1,1))
plot(x1,y1,type="p",col=adjustcolor(4,0.05),ylab="Species Richness",xlab="Area (ha)",las=1, bty="l",pch=16, cex=2, axes=F, ylim=c(0,70))
box(bty="l")
axis(2,las=1)
axis(1,at=1:10)
dev.off()

```

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=2cm,xshift=-3cm] at (current page.south east)
{\includegraphics[height=1.5in]{specareaA.pdf}};
\end{tikzpicture}

## Correlations

Interested in the relationship between two samples

- Dependent variable:
    - data we are interested in explaining
    - Y-axis
    - *response* variable
    
- Independent variable:
    - data used to describe variation in dependent variable 
    - X-axis
    - *explanatory* variable

```{r,eval=TRUE,echo=FALSE,results="hide"}
pdf("Images//Corr//specareaB.pdf",height=3,width=4)
par(mfrow=c(1,1),oma=c(0,0,0,0),mar=c(4,4,1,1),mgp=c(1,1,1))
plot(AR,SR,type="p",bg=adjustcolor(4,0.4),xlab="Independent variable",ylab="Dependent variable",ylim=c(5,30),las=1, bty="l",pch=21, cex=2,axes=F)
box(bty="l")
dev.off()
```

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=2cm,xshift=-3cm] at (current page.south east)
{\includegraphics[height=1.5in]{specareaB.pdf}};
\end{tikzpicture}

## Correlations

Interested in the relationship between two samples

- Dependent variable:
    - data we are interested in explaining
    - Y-axis
    - *response* variable
    
- Independent variable:
    - data used to describe variation in dependent variable 
    - X-axis
    - *explanatory* variable

- Dealing with *pairs* of values!

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=2cm,xshift=-3cm] at (current page.south east)
{\includegraphics[height=1.5in]{specareaB.pdf}};
\end{tikzpicture}

## Correlations

What is the sign of the correlation?

```{r,eval=TRUE,echo=FALSE,results="hide"}
X <- runif(25,0,10)
Pos <- rnorm(25,10 + 2*X,2)
No <- rnorm(25,10 + 0*X,2)
Neg <- rnorm(25,10 - 2*X,2)
pdf("Images//Corr//signA.pdf",height=2,width=6)
par(mfrow=c(1,3),oma=c(0,0,0,0),mar=c(4,4,3,1),mgp=c(1,1,1))
plot(X,Pos,type="p",bg=adjustcolor(1,0.3),xlab="Independent",ylab="Dependent",las=1, bty="l",pch=21, cex=1.8,axes=F)
box(bty="l")
plot(X,No, type="p",bg=adjustcolor(1,0.3),xlab="Independent",ylab="Dependent",las=1, bty="l",pch=21, cex=1.8,axes=F)
box(bty="l")
plot(X,Neg,type="p",bg=adjustcolor(1,0.3),xlab="Independent",ylab="Dependent",las=1, bty="l",pch=21, cex=1.8,axes=F)
box(bty="l")
dev.off()
pdf("Images//Corr//signB.pdf",height=2,width=6)
par(mfrow=c(1,3),oma=c(0,0,0,0),mar=c(4,4,3,1),mgp=c(1,1,1))
plot(X,Pos,type="p",bg=adjustcolor(3,0.3),xlab="Independent",ylab="Dependent",las=1, bty="l",pch=21, cex=1.8, main="Positive Correlation",axes=F)
box(bty="l")
plot(X,No, type="p",bg=adjustcolor(1,0.3),xlab="Independent",ylab="Dependent",las=1, bty="l",pch=21, cex=1.8, main="No Correlation",axes=F)
box(bty="l")
plot(X,Neg,type="p",bg=adjustcolor(2,0.3),xlab="Independent",ylab="Dependent",las=1, bty="l",pch=21, cex=1.8, main="Negative Correlation",axes=F)
box(bty="l")
dev.off()
```

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1cm] at (current page)
{\includegraphics[height=1.5in]{signA.pdf}};
\end{tikzpicture}

## Correlations

What is the sign of the correlation?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1cm] at (current page)
{\includegraphics[height=1.5in]{signB.pdf}};
\end{tikzpicture}

## Correlations

Lets play a game!

\vspace{1cm}
\large
\begin{center}
\textcolor{blue}{http://www.istics.net/Correlations/}
\end{center}

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=3cm] at (current page.south)
{\includegraphics[height=1.5in]{jigsaw.jpg}};
\end{tikzpicture}

## Correlations

We need a way to 

- quantify correlations/relationships
- assess whether correlations/relationships are *significant*

## Correlations

We need a *test*!

1. Spearman's rank test ($r_s$)

    - determine the strength of the link between 2 samples 
    - data are *not* normally distributed
    - realtionship not linear
        - but still exhibits a positive of negative trend
        - is not u-shaped or n-shaped
    - use the *ranks* of values
    - correlation strenght ranges from -1 to 1
        - -1: perfect *negative* correlation
        - 1: perfect *positive* correlation
        - 0: no correlation
        
## Correlations

We need a *test*!

2. Pearson's product moment  ($r$)

    - determine the strength of the link between 2 samples 
    - data *are* normally distributed
    - relationship assumed to be linear
        - positive of negative trend
        - not u- or n-shaped
    - use actual values
    - correlation strenght ranges from -1 to 1
        - -1: perfect *negative* correlation
        - 1: perfect *positive* correlation
        - 0: no correlation

## Correlations

1. Spearman's rank test ($r_s$)

    - determine the strength of the link between 2 samples 
    - data are *not* normally distributed
    - not neccesarily linear
    - use the *ranks* of values

2. Pearson's product moment  ($r$)

    - determine the strength of the link between 2 samples 
    - data *are* normally distributed
    - assumed to be linear linear
    - use actual values

- For both, the *correlation coefficient* ranges from -1 to 1
    - -1: perfect *negative* correlation
    - 1: perfect *positive* correlation
    - 0: no correlation

## Example: the Mayfly data

```{r,echo=FALSE, results="hide"}
mayfly <- data.frame(Speed = c(2,3,5,9,14,24,29,34),
                     Abundance = c(6,3,5,23,16,12,48,43))
pdf("Images//Corr//spearMayfly.pdf",height=3.5,width=3.5, bg="white")
par(mfrow=c(1,1),oma=c(0,0,0,0),mar=c(4,4,3,1))
with(mayfly,
     plot(Speed, Abundance,pch=21,bg=adjustcolor(4,0.5),cex=1.5, las=1,ylim=c(0,50))
)
dev.off()
```

\vfill

```{r}
head(mayfly)
```

\vfill

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-4cm] at (current page.east)
{\includegraphics[height=3in]{spearMayfly.pdf}};
\end{tikzpicture}

## Spearman's rank test ($r_s$) - the hypothesis

Before conducting any statistical test, we need to state the hypotheses!

- $H_0$ (the *null* hypothesis):

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=2cm,xshift=-2cm] at (current page.south east)
{\includegraphics[height=2in]{QM.png}};
\end{tikzpicture}

## Spearman's rank test ($r_s$) - the hypothesis

Before conducting any statistical test, we need to state the hypotheses!

- $H_0$ (the *null* hypothesis):

*There is no correlation between stream speed and mayfly abundance*

## Spearman's rank test ($r_s$) - the hypothesis

Before conducting any statistical test, we need to state the hypotheses!

- $H_0$ (the *null* hypothesis):

*There is no correlation between stream speed and mayfly abundance*

- $H_1$ (the *alternative* hypothesis):

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=2cm,xshift=-2cm] at (current page.south east)
{\includegraphics[height=2in]{QM.png}};
\end{tikzpicture}

## Spearman's rank test ($r_s$) - the hypothesis

Before conducting any statistical test, we need to state the hypotheses!

- $H_0$ (the *null* hypothesis):

*There is no correlation between stream speed and mayfly abundance*

- $H_1$ (the *alternative* hypothesis):

*There is a positive correlation between stream speed and mayfly abundance*

## Spearman's rank test ($r_s$) - the statistical test

$$r_s = 1 - \frac{6 \times \sum D^2}{n(n^2 - 1)}$$

- programs (`Excel` and `R`) will do the math for us
- \underline{BUT} we should be aware of what's going on!

## Spearman's rank test ($r_s$) - the statistical test

$$r_s = 1 - \frac{6 \times \sum D^2}{n(n^2 - 1)}$$

- programs (`Excel` and `R`) will do the math for us
- \underline{BUT} we should be aware of what's going on!

```{r}
mayfly
```

## Spearman's rank test ($r_s$) - the statistical test

$$r_s = 1 - \frac{6 \times \sum D^2}{n(n^2 - 1)}$$

- first, calculate the *ranks* of the values: speed

```{r,echo=FALSE}
mayfly$Speed.rank <- order(mayfly$Speed)
```

```{r}
mayfly
```

## Spearman's rank test ($r_s$) - the statistical test

$$r_s = 1 - \frac{6 \times \sum D^2}{n(n^2 - 1)}$$

- first, calculate the *ranks* of the values: abundance

```{r,echo=FALSE}
mayfly$Abundance.rank <- rank(mayfly$Abundance)
```

```{r}
mayfly
```

## Spearman's rank test ($r_s$) - the statistical test

$$r_s = 1 - \frac{6 \times \sum D^2}{n(n^2 - 1)}$$

- then, calculate the difference in the *ranks*: $D$

```{r,echo=FALSE}
mayfly$Diff <- mayfly$Abundance.rank - mayfly$Speed.rank
```

```{r}
mayfly
```

## Spearman's rank test ($r_s$) - the statistical test

$$r_s = 1 - \frac{6 \times \sum D^2}{n(n^2 - 1)}$$

- then, square the difference in the *ranks*: $D^2$

```{r,echo=FALSE}
mayfly$Diff.sq <- mayfly$Diff^2
```

```{r}
mayfly
```

## Spearman's rank test ($r_s$) - the statistical test

$$r_s = 1 - \frac{6 \times \sum D^2}{n(n^2 - 1)}$$

- so now we have all the pieces!
    - $n$: number of pairs of observations (each has an $x$ and a $y$)
    - $D$: difference in ranks between variables
    - $D^2$: the difference squared

## Spearman's rank test ($r_s$) - in practice

$$r_s = 1 - \frac{6 \times \sum D^2}{n(n^2 - 1)}$$

```{r}
mayfly
```

## Spearman's rank test ($r_s$) - in practice

$$r_s = 1 - \frac{6 \times \sum D^2}{n(n^2 - 1)}$$

$$r_s = 0.81$$

What does this tell us?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=2cm,xshift=-2cm] at (current page.south east)
{\includegraphics[height=2in]{QM.png}};
\end{tikzpicture}

## Spearman's rank test ($r_s$) - in practice

$$r_s = 1 - \frac{6 \times \sum D^2}{n(n^2 - 1)}$$

$$r_s = 0.81$$

Two useful pieces of information:

1. Sign of the correlation
    - positive value means *positive* correlation
    
## Spearman's rank test ($r_s$) - in practice

$$r_s = 1 - \frac{6 \times \sum D^2}{n(n^2 - 1)}$$

$$r_s = 0.81$$

Two useful pieces of information:

1. Sign of the correlation
    - positive value means *positive* correlation
2. Strength of the correlation
    - $r$ ranges from -1 to 1
    - $r_s = 0.81$ is *strongly* positive
    - but is this *significant*?
    
    
## Spearman's rank test ($r_s$) - conclusions

As with many statistical tests, determine significance based on:

- *significance level*
- *sample size*
- *critical value*

## Spearman's rank test ($r_s$) - conclusions

As with many statistical tests, determine significance based on:

- *significance level*
- *sample size*
- *critical value*

If the test statistic is \underline{greater than} the critical value:

- *reject* the null hypothesis
- there *is* a statistically significant positive correlation

## Spearman's rank test ($r_s$) - conclusions

As with many statistical tests, determine significance based on:

- *significance level*
- *sample size*
- *critical value*

If the test statistic is \underline{greater than} the critical value:

- *reject* the null hypothesis
- there *is* a statistically significant positive correlation

If the test statistic is \underline{less than} the critical value:

- *fail to reject* the null hypothesis
- there is no evidence of a statistically significant positive correlation

## Spearman's rank test ($r_s$) - mayfly conclusions

- *significance level* (typically 5%)
- *sample size* (number of pairs = 8)
- *critical value* (0.738)
- *test stasistic* ($r_s = 0.81$)
- the *test stasistic* is \textcolor{blue}{??} than the *critical value*

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=3cm] at (current page.south)
{\includegraphics[height=2in]{spearTab.jpg}};
\end{tikzpicture}

## Spearman's rank test ($r_s$) - mayfly conclusions

So, $r_s > 0.738$:

- \textcolor{blue}{reject} the null hypothesis
- \textcolor{blue}{accept} the alternative hypothesis

There *is* a statistically significant positive correlation between stream flow and mayfly abundance!


## Pearson's product moment  ($r$)

2. Pearson's product moment  ($r$)

## Pearson's product moment  ($r$)

2. Pearson's product moment  ($r$)

    - determine the strength of the link between 2 samples 
    - data *are* normally distributed
    - relationship assumed to be linear
    - use actual values

## Pearson's product moment  ($r$)

2. Pearson's product moment  ($r$)

    - determine the strength of the link between 2 samples 
    - data *are* normally distributed
    - relationship assumed to be linear
    - use actual values
    - -1: perfect *negative* correlation
    - 1: perfect *positive* correlation
    - 0: no correlation

## Pearson's product moment  ($r$)

2. Pearson's product moment  ($r$)

    - determine the strength of the link between 2 samples 
    - data *are* normally distributed
    - \textcolor{blue}{relationship assumed to be linear}
    - use actual values
    - -1: perfect *negative* correlation
    - 1: perfect *positive* correlation
    - 0: no correlation

## Pearson's product moment  ($r$)

The linear assumption:

- relationship between two variables can be described by the equation of a straight line:

$$y = mx + c$$

- $y$: the dependent/response variable
- $x$: the independent/explanatory variable
- $m$: the *slope* of the relationship
- $c$: the *intercept*

## Pearson's product moment  ($r$) - an example

Relationship between area \& species richness (by hand/eye):

$$species = m \times area + c$$

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=3cm] at (current page.south)
{\includegraphics[height=2in]{specareaWith.pdf}};
\end{tikzpicture}

## Pearson's product moment  ($r$) - an example

Relationship between area \& species richness (by hand/eye):

$$species = m \times area + c$$

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=3cm] at (current page.south)
{\includegraphics[height=2in]{specareaWithout.pdf}};
\end{tikzpicture}

## Pearson's product moment  ($r$)

2. Pearson's product moment  ($r$)

- calculates the strength of the correlation
    - similar to Spearman's coefficient
- *also* the equation of the best fit line
    - calculates the slope
    - calculates the intercepts

\begin{center}    
Pearson's product moment calculates strength of correlation \& parameters of linear relationship between two variables.
\end{center}    

## Example: the Mayfly data

```{r,echo=FALSE, results="hide"}
mayfly <- data.frame(Speed = c(2,3,5,9,14,24,29,34),
                     Abundance = c(6,3,5,23,16,12,48,43))
pdf("Images//Corr//spearMayfly.pdf",height=3.5,width=3.5, bg="white")
par(mfrow=c(1,1),oma=c(0,0,0,0),mar=c(4,4,3,1))
with(mayfly,
     plot(Speed, Abundance,pch=21,bg=adjustcolor(4,0.5),cex=1.5, las=1,ylim=c(0,50))
)
dev.off()
```

\vfill

```{r}
head(mayfly)
```

\vfill

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-4cm] at (current page.east)
{\includegraphics[height=3in]{spearMayfly.pdf}};
\end{tikzpicture}

## Pearson's product moment  ($r$) - the hypothesis

Before conducting any statistical test, we need to state the hypotheses!

- $H_0$ (the *null* hypothesis):

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=2cm,xshift=-2cm] at (current page.south east)
{\includegraphics[height=2in]{QM.png}};
\end{tikzpicture}

## Pearson's product moment  ($r$) - the hypothesis

Before conducting any statistical test, we need to state the hypotheses!

- $H_0$ (the *null* hypothesis):

*There is no correlation between stream speed and mayfly abundance*

## Pearson's product moment  ($r$) - the hypothesis

Before conducting any statistical test, we need to state the hypotheses!

- $H_0$ (the *null* hypothesis):

*There is no correlation between stream speed and mayfly abundance*

- $H_1$ (the *alternative* hypothesis):

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=2cm,xshift=-2cm] at (current page.south east)
{\includegraphics[height=2in]{QM.png}};
\end{tikzpicture}

## Pearson's product moment  ($r$) - the hypothesis

Before conducting any statistical test, we need to state the hypotheses!

- $H_0$ (the *null* hypothesis):

*There is no correlation between stream speed and mayfly abundance*

- $H_1$ (the *alternative* hypothesis):

*There is a positive correlation between stream speed and mayfly abundance*

## Pearson's product moment  ($r$) - the hypothesis

Before conducting any statistical test, we need to state the hypotheses!

- $H_0$ (the *null* hypothesis):

*There is no correlation between stream speed and mayfly abundance*

- $H_1$ (the *alternative* hypothesis):

*There is a positive correlation between stream speed and mayfly abundance*

\vspace{0.5cm}

NOTE: the hypotheses remain the same, but the assumption about the data are different. The assumptions determine the appropriate model!

## Pearson's product moment  ($r$) - the straight line

Equation of a straight line: 

$$y = mx + c$$

## Pearson's product moment  ($r$) - the slope

Equation of a straight line: 

$$y = \textcolor{blue}{m}x + c$$

Calculating the slope: 

$$\textcolor{blue}{m} = \frac{\sum(x-\bar{x})(y-\bar{y})}{\sum(x-\bar{x})^2}$$

## Pearson's product moment  ($r$) - the intercept

Equation of a straight line: 

$$y = mx + \textcolor{blue}{c}$$

Calculating the slope: 

$$m = \frac{\sum(x-\bar{x})(y-\bar{y})}{\sum(x-\bar{x})^2}$$

Calculating the intercept: 

$$\textcolor{blue}{c} = \bar{y} - m\bar{x}$$

## Pearson's product moment  ($r$) - the linear equation

Equation of a straight line: 

$$y = mx + \textcolor{blue}{c}$$

- programs (`Excel` and `R`) will do the math for us
- \underline{BUT} we should be aware of what's going on!
    - calculating the *slope* from the data
    - calculating the *intercept* from the data
    
## Pearson's product moment  ($r$) - the statistical test

Equation of a straight line: $y = mx + c$

- but, is the linear relationship *significant*?
- we can calculate Pearson's correlation coefficient

$$r=\frac{\sum (x-\bar{x})(y-\bar{y})}  {\sqrt{\sum (x-\bar{x})^2 \sum (y-\bar{y})^2}}$$

## Pearson's product moment  ($r$) - the statistical test

Equation of a straight line: $y = mx + c$

- but, is the linear relationship *significant*?
- we can calculate Pearson's correlation coefficient

$$r=\frac{\sum (x-\bar{x})(y-\bar{y})}  {\sqrt{\sum (x-\bar{x})^2 \sum (y-\bar{y})^2}}$$

```{r,echo=FALSE}
mayfly <- mayfly[,1:2]
```

```{r}
mayfly
```

## Pearson's product moment  ($r$) - the statistical test

Equation of a straight line: $y = mx + c$

- but, is the linear relationship *significant*?
- we can calculate Pearson's correlation coefficient

$$r=\frac{\sum (x-\bar{x})(y-\bar{y})}  {\sqrt{\sum (x-\bar{x})^2 \sum (y-\bar{y})^2}}$$

$$r = 0.84$$


## Pearson's product moment  ($r$) - in practice

$$r=\frac{\sum (x-\bar{x})(y-\bar{y})}  {\sqrt{\sum (x-\bar{x})^2 \sum (y-\bar{y})^2}}$$

$$r = 0.84$$

Again we have two useful pieces of information:

1. Sign of the correlation
    - positive value means *positive* correlation
2. Strength of the correlation
    - $r$ ranges from -1 to 1
    - $r_s = 0.84$ is *strongly* positive
    - but is this *significant*?

## Pearson's product moment  ($r$) - conclusions

As with many statistical tests, determine significance based on:

- *significance level*
- *sample size*
- *critical value*

## Pearson's product moment  ($r$) - conclusions

As with many statistical tests, determine significance based on:

- *significance level*
- *sample size*
- *critical value*

If the test statistic is \underline{greater than} the critical value:

- *reject* the null hypothesis
- there *is* a statistically significant positive correlation

## Pearson's product moment  ($r$) - conclusions

As with many statistical tests, determine significance based on:

- *significance level*
- *sample size*
- *critical value*

If the test statistic is \underline{greater than} the critical value:

- *reject* the null hypothesis
- there *is* a statistically significant positive correlation

If the test statistic is \underline{less than} the critical value:

- *fail to reject* the null hypothesis
- there is no evidence of a statistically significant positive correlation

## Pearson's product moment  ($r$) - mayfly conclusions

- *significance level* (typically 5%)
- *sample size* (df = number of pairs - 2 = 8 - 2 = 6)
- *critical value* (0.707)
- *test stasistic* ($r = 0.84$)
- the *test stasistic* is \textcolor{blue}{??} than the *critical value*

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=3cm] at (current page.south)
{\includegraphics[height=2in]{pearsTab.jpg}};
\end{tikzpicture}

## Pearson's product moment  ($r$) - mayfly conclusions

So, $r > 0.707$:

- \textcolor{blue}{reject} the null hypothesis
- \textcolor{blue}{accept} the alternative hypothesis

There *is* a statistically significant positive correlation between stream flow and mayfly abundance!

## Mayfly conclusions

Spearman's rank correlation coefficient:

- $r_s > 0.738$:
- \textcolor{blue}{reject} the null hypothesis
- \textcolor{blue}{accept} the alternative hypothesis

Pearson's correlation coefficient:

- $r > 0.707$:
- \textcolor{blue}{reject} the null hypothesis
- \textcolor{blue}{accept} the alternative hypothesis

Conclusion:

- same regardless of assumptions!  
- there *is* a statistically significant positive correlation between stream flow and mayfly abundance!

## Correlations in practice

- we *could* do it by hand (UGH!)
- we *prefer* to use built in functions
    - in `Excel`
    - in `R`

## Example: the Mayfly data

```{r,echo=FALSE, results="hide"}
mayfly <- data.frame(Speed = c(2,3,5,9,14,24,29,34),
                     Abundance = c(6,3,5,23,16,12,48,43))
pdf("Images//Corr//spearMayfly.pdf",height=3.5,width=3.5, bg="white")
par(mfrow=c(1,1),oma=c(0,0,0,0),mar=c(4,4,3,1))
with(mayfly,
     plot(Speed, Abundance,pch=21,bg=adjustcolor(4,0.5),cex=1.5, las=1,ylim=c(0,50))
)
dev.off()
```

\vfill

```{r}
head(mayfly)
```

\normalsize

Spearman: $r_s = 0.81$  
Pearson: $r = 0.84$

\vfill

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-4cm] at (current page.east)
{\includegraphics[height=3in]{spearMayfly.pdf}};
\end{tikzpicture}

## Spearman's rank test in \texttt{Excel}

$$r_s = 1 - \frac{6 \times \sum D^2}{n(n^2 - 1)}$$

In \texttt{Excel} 

- an automated routine doe snot exist
- have to do it by hand
- :(

## Pearson's product moment in \texttt{Excel}

Pearson's correlation coefficient:

- $r=\frac{\sum (x-\bar{x})(y-\bar{y})}  {\sqrt{\sum (x-\bar{x})^2 \sum (y-\bar{y})^2}}$

- `CORREL(response, predictor)` 

Slope of the best fit line:

- $m = \frac{\sum(x-\bar{x})(y-\bar{y})}{\sum(x-\bar{x})^2}$

- `SLOPE(response, predictor)`

Intercept of the best fit line:

- $c = \bar{y} - m\bar{x}$

- `INTERCEPT(response, predictor)`

## Pearson's product moment in \texttt{Excel}

Pearson's correlation coefficient:

- $r=\frac{\sum (x-\bar{x})(y-\bar{y})}  {\sqrt{\sum (x-\bar{x})^2 \sum (y-\bar{y})^2}}$

- Analysis toolpack - `Correlation`

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1.5cm,xshift=-2.6cm] at (current page)
{\includegraphics[height=1.2in]{atpA.png}};
\node[yshift=-1.5cm,xshift=3cm] at (current page)
{\includegraphics[height=1.2in]{atpB.png}};
\end{tikzpicture}

## Spearman's rank test in \texttt{R}

```{r}
#make the mayfly 'data.frame' by hand:
mayfly <- data.frame(Speed = c(2,3,5,9,14,24,29,34),
                     Abundance = c(6,3,5,23,16,12,48,43))
```

## Spearman's rank test in \texttt{R}

```{r}
#make the mayfly 'data.frame' by hand:
mayfly <- data.frame(Speed = c(2,3,5,9,14,24,29,34),
                     Abundance = c(6,3,5,23,16,12,48,43))
mayfly
```

## Spearman's rank test in \texttt{R}

$$r_s = 1 - \frac{6 \times \sum D^2}{n(n^2 - 1)}$$

- Spearman correlation coefficient ($r_s$)

```{r,eval=FALSE}
cor(var1, var2, method='spearman')
```

\normalsize

- Spearman correlation coefficient *and* significance test

```{r,eval=FALSE}
cor.test(var1, var2, method='spearman')
```

## Spearman's rank test in \texttt{R}

- Spearman correlation coefficient ($r_s$)

```{r}
cor(mayfly$Speed, mayfly$Abundance, method='spearman')
```

## Spearman's rank test in \texttt{R}

- Spearman correlation coefficient ($r_s$)

```{r}
cor(mayfly$Speed, mayfly$Abundance, method='spearman')
```

\normalsize

- Spearman correlation coefficient *and* significance test

```{r}
cor.test(mayfly$Speed, mayfly$Abundance, method='spearman')
```

## Pearson's product moment in \texttt{R}

- $r=\frac{\sum (x-\bar{x})(y-\bar{y})}  {\sqrt{\sum (x-\bar{x})^2 \sum (y-\bar{y})^2}}$

- Pearson's correlation coefficient ($r$)

```{r,eval=FALSE}
cor(var1, var2, method='pearson')
```

\normalsize

- Pearson's correlation coefficient *and* significance test

```{r,eval=FALSE}
cor.test(var1, var2, method='pearson')
```

## Pearson's product moment in \texttt{R}

- Pearson's correlation coefficient ($r$)

```{r,eval=FALSE}
cor(mayfly$Speed, mayfly$Abundance, method='pearson')
```

## Pearson's product moment in \texttt{R}

- Pearson's correlation coefficient ($r$)

```{r}
cor(mayfly$Speed, mayfly$Abundance, method='pearson')
```

## Pearson's product moment in \texttt{R}

- Pearson's correlation coefficient ($r$)

```{r}
cor(mayfly$Speed, mayfly$Abundance, method='pearson')
```

\normalsize

- Pearson's correlation coefficient *and* significance test

```{r,eval=FALSE}
cor.test(mayfly$Speed, mayfly$Abundance, method='pearson')
```

## Pearson's product moment in \texttt{R}

- Pearson's correlation coefficient ($r$)

```{r}
cor(mayfly$Speed, mayfly$Abundance, method='pearson')
```

\normalsize

- Pearson's correlation coefficient *and* significance test

```{r}
cor.test(mayfly$Speed, mayfly$Abundance, method='pearson')
```

## Linear equation in \texttt{R}

To calculate the slope and the intercept of the best fit line:

- use a *linear model*
- in `R` use the function `lm(respnse ~ predictor)`

## Linear equation in \texttt{R}

To calculate the slope and the intercept of the best fit line:

- use a *linear model*
- in `R` use the function `lm(respnse ~ predictor)`

```{r}
lm(mayfly$Abundance ~ mayfly$Speed)
```

## Linear equation in \texttt{R}

To calculate the slope and the intercept of the best fit line:

- use a *linear model*
- in `R` use the function `lm(respnse ~ predictor)`

```{r, my.size=TRUE, size="\\tiny"}
summary(lm(mayfly$Abundance ~ mayfly$Speed))
```

## Its all related!

```{r, my.size=TRUE, size="\\tiny"}
cor.test(mayfly$Speed, mayfly$Abundance, method='pearson')
summary(lm(mayfly$Abundance ~ mayfly$Speed))
```

## Its all related!

```{r, my.size=TRUE, size="\\tiny"}
cor.test(mayfly$Speed, mayfly$Abundance, method='spearman')
summary(lm(mayfly$Abundance ~ mayfly$Speed))
```

## Group exercise: salamander length correlations

Correlations in `R`: 

1. Read in the data
2. Make a plot of 'total length' in response to 'SVL' length
3. Decide whether to use the Pearson or Spearman test

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-2.75cm,xshift=4.2cm] at (current page)
{\includegraphics[height=1.5in]{mander.png}};
\end{tikzpicture}

## Group exercise: salamander length correlations

Correlations in `R`: 

- Read in the data
- Make a plot of 'total length' in response to 'SVL' length
- Decide whether to use the Pearson or Spearman test
- Conduct the analysis:
    - what is the correlation coefficient?
    - what is the intercept of the relationship?
    - what is the slope of the relationship?
    - is the relationship significant?
    
\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-2.75cm,xshift=4.2cm] at (current page)
{\includegraphics[height=1.5in]{mander.png}};
\end{tikzpicture}

## Group exercise: salamander length correlations

Correlations in `Excel`: 

- Create an Excel worksheet that show the same values you got in `R`:
    - the correlation coefficient
    - the intercept of the relationship
    - the slope of the relationship
    - the $p$-value

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-2.75cm,xshift=4.2cm] at (current page)
{\includegraphics[height=1.5in]{mander.png}};
\end{tikzpicture}

## Group exercise: salamander length correlations

Correlations in `Excel`: 

- Create an Excel worksheet that show the same values you got in `R`:
    - the correlation coefficient
    - the intercept of the relationship
    - the slope of the relationship
    - the $p$-value (see below)
- **Submit the workbook**

To compute the $p$-value, you need to use `T.DIST.2T()` in Excel. This needs $t$,  *test statistic*, which you can calculate that using $r$ (the correlation coefficient):

$$t = \frac{r \times \sqrt{n-2}}{\sqrt{1-r^2}}$$
