---
title: "Graphical Data Exploration: pre-class assignment"
subtitle: "NRC290B Introduction to Quantitative Ecology"
author: "Michael France Nelson"
date: "Spring 2020"
output: 
  html_document:
    css: !expr here::here("css", "styles.css")
    number_sections: TRUE
    toc: true
    toc_float: true
---






```{r setup, include=FALSE}
require(here)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(root.dir = here(), error = TRUE)
knitr::opts_knit$set(echo = TRUE, error = TRUE, root.dir = here())
source(here("data", "environment_vars.R"))

due_date = ISOdatetime(2020, 3, 1, 23, 55, 0)
format(due_date, due_date_format)


```


```{r anova answer key, echo = FALSE, include=FALSE}



# sample format
upper  = c(23, 25, 27, 28, 19, 26, 27)
middle = c(32, 37, 31, 28, 26, 29    )
lower  = c(24, 26, 27, 28, 25        )

# recording format
sward_heights = 
  rbind(
    data.frame(site = "upper",  height = upper),
    data.frame(site = "middle", height = middle),
    data.frame(site = "lower",  height = lower)
  )


all    = c(upper, middle, lower)

n_upper  = length(upper)   # B9
n_middle = length(middle)  # C9
n_lower  = length(lower)   # D9

df_upper  = n_upper - 1
df_middle = n_middle - 1
df_lower  = n_lower - 1

df_within = df_upper + df_middle + df_lower

n_between = 3
df_between = n_between - 1

n_total = sum(n_upper, n_middle, n_lower) # A9
df_total = n_total - 1

df = c(between = df_between,  within = df_within, total = df_total)

mean_upper  = mean(upper)  # B10
mean_middle = mean(middle) # C10
mean_lower  = mean(lower)  # D10

ss_upper  = var(upper)  * (n_upper  - 1) # B11
ss_middle = var(middle) * (n_middle - 1) # C11
ss_lower  = var(lower)  * (n_lower  - 1) # D11

grand_mean = mean(c(upper, middle, lower)) # C12

ss_within  = sum(ss_upper, ss_middle, ss_lower) # A13

ss_between_upper  = ((mean_upper  - grand_mean) ^ 2) * n_upper  # B14
ss_between_middle = ((mean_middle - grand_mean) ^ 2) * n_middle # C14
ss_between_lower  = ((mean_lower  - grand_mean) ^ 2) * n_lower  # D14

ss_between = sum(ss_between_upper, ss_between_middle, ss_between_lower)

ss_total = var(all) * (n_total - 1)

ss_total == (ss_between + ss_within)

source = c(between = ss_between, within = ss_within, total = ss_total)
df     = c(between = df_between,  within = df_within, total = df_total)
ms     = c(source / df)[1:2]

f_val = ms[1] / ms[2]
p_crit = 0.05

p_obs = pf(f_val, df_between, df_within, lower.tail = FALSE)
f_crit = qf(p_crit, df_between, df_within, lower.tail = FALSE)


# post hoc analysis
fit1 = aov(height ~ site, data = sward_heights)
summary(fit1)
TukeyHSD(fit1, "site", ordered = TRUE)
```

# Overview

This is an adaptation of the Excel exercise on pages 267 - 268 from Gardner: Statistics for Ecologists using R and Excel




# The data

This dataset includes measurements of grass height in three sites (swards).

Normally you would record data in the *recording format* listed below.

For the sake of this exercise, you will need both the sample and recording formats.

```{r build data}

# sample format
upper  = c(23, 25, 27, 28, 19, 26, 27)
middle = c(32, 37, 31, 28, 26, 29    )
lower  = c(24, 26, 27, 28, 25        )

# recording format
sward_heights = 
  rbind(
    data.frame(site = "upper",  height = upper),
    data.frame(site = "middle", height = middle),
    data.frame(site = "lower",  height = lower)
  )

# view a sample of the recording format:
head(sward_heights)
```



# Performing an Analysis of Variance by hand

Typically, you would have software do the ANOVA calculations for you, but it is helpful to do an ANOVA by hand learn how the analysis works.

Here are some code templates for performing the ANOVA and building the table.


You will need to calculate some intermediate quantities to do the analysis:

* The total number of observations
* The number of groups
* The number of observations in each group
* Within-group, between-group, and total degrees of freedom


## Calculating the sample numbers

To calculate the numbers of within-group samples use the `length()` function.
Fill in the missing code
```{r, eval = FALSE}
n_upper  = length(upper)   # B9
n_middle =
n_lower  =
```




The number of groups:
```{r, eval = FALSE}
n_between = 3
```

Total number of observations:

```{r, eval = FALSE}
n_total = sum(n_upper, n_middle, n_lower)
```

## Degrees of freedom

### Within group DF

The degrees of freedom for each groups are just the counts minus 1:

```{r, eval = FALSE}

df_upper  = n_upper - 1
df_middle =
df_lower  =

```
The total within-group degrees of freedom is just the sum of the individual groups' degrees of freedom:
```{r}
df_within = df_upper + df_middle + df_lower
```

You should get:
```{r}
df_within
```


### Total DF

Total degrees of freedom is the total number of observations minus 1:

```{r, eval = FALSE}
df_total  =
```

You should get:
```{r}
df_total
```

### Vector of DFs

Create a vector to hold all of the degrees of freedom
```{r, eval = FALSE}
df = c(between = df_between,  within = df_within, total = df_total)
```

Your degrees of freedom vector should be:
```{r}
df
```


## Calculating means

Now we need the different means to calculate the Sums of Squares for the analysis.


### Grand mean
The grand mean is the mean of every observation.  Use the functions `c()` and `mean()` to find it:

```{r}
grand_mean = mean(c(upper, middle, lower)) # C12
```


### Within-group means

The within-group means calculations are similar

```{r, eval = FALSE}
mean_upper  = mean(upper)  # B10
mean_middle =              # C10
mean_lower  =              # D10
```


## Sums of Squares

### Within-group SS
We can use the function `var()` to calculate the within-group sum of squares:
```{r, eval = FALSE}
ss_upper  = var(upper)  * (n_upper  - 1) # B11
ss_middle =                              # C11
ss_lower  =                              # D11
```

You should get:
```{r}
ss_upper
ss_middle
ss_lower
```


The total within-group SS:

```{r, eval = FALSE}
ss_within  = sum(ss_upper, ss_middle, ss_lower) # A13
```

You should get: 
```{r}
ss_within
```


### Between-group SS
The between-group SS terms:
```{r, eval = FALSE}
ss_between_upper  = ((mean_upper  - grand_mean) ^ 2) * n_upper  # B14
ss_between_middle =                                             # C14
ss_between_lower  =                                             # D14
  
ss_between = sum(ss_between_upper, ss_between_middle, ss_between_lower)
```

You should get
```{r}
ss_between_upper
ss_between_middle
ss_between_lower

ss_between
```



### Total SS

You can calculate the total SS in two ways.  They should produce the same quantity

```{r, eval = FALSE}
all    = c(upper, middle, lower)
ss_total_1 = var(all) * (n_total - 1)
ss_total_2 = (ss_between + ss_within)

# This should return true if your calculations above were correct
ss_total_1 == ss_total_2

```

We'll go with the first way:
```{r, eval = FALSE}
var(all) * (n_total - 1)
```

You should get
```{r}
ss_total
```

### Vector of SS

Make a vector of SS for each source
```{r, eval = FALSE}
source = c(between = ss_between, within = ss_within, total = ss_total)
```

You should get:
```{r}
source
```



## ANOVA table components

Now we can assemble our ANOVA table


### SS and DF

We already calculated our degrees of freedom and sums of squares.

### Mean Squares

Calculate the mean squares for the between and within terms
```{r, eval = FALSE}
ms     = c(source / df)[1:2]
```

We don't calculate a mean squares for the total.

You should get:
```{r}
ms
```



## Significance


### F
The F-value is just the ratio of between to within group MS

```{r, eval = FALSE}
f_val = ms[1] / ms[2]
```

You should get:
```{r}
f_val
```



### P

We need to decide our critical p-value.  We'll use 5%.
```{r}
p_crit = 0.05
```



### Critical F

We use the `qf()` function to find a critical F value.  Be sure to use `lower.tail = FALSE`

```{r}
f_crit = qf(p_crit, df_between, df_within, lower.tail = FALSE)
```

You should get:
```{r}
f_crit
```


### Observed p-value

We calculate an exact p-value and compare it to our critical p-value:
```{r, eval = FALSE}
p_obs = pf(f_val, df_between, df_within, lower.tail = FALSE)
```

You should get:
```{r}
p_obs
```





# Complete ANOVA table

It's a pain to combine our quantities into a table in R, but we could paste them into a spreadsheet.

Just to verify that your calculations were correct, you should have gotten

```{r}
# sums of squares
source

# degrees of freedom
df

# mean squares
ms

# F-value
f_val

# p-value
p_obs

# Critical F-value
f_crit

```


### R's `aov()`

Compare your calculations to R's Analysis of Variance function:
```{r}
fit1 = aov(height ~ site, data = sward_heights)
summary(fit1)
```






# Post-hoc analysis

We won't do a post-hoc Tukey analysis by hand.

R's calculations produce slightly different results from the book's Excel calculations.

```{r}
# post hoc analysis
TukeyHSD(fit1, "site", ordered = TRUE)
```



# Assignment

Complete the code in the provided template script.  You can check your calculations on this page.

