---
title: "Week 5: Exploring Data"
subtitle: "Session 1"
date: Spring 2020
output:
  beamer_presentation:
    pandoc_args: !expr paste0(here::here('css', 'beamer.yaml'))
header-includes:
  \input{`r here::here("css", "headers_tikz.tex")`}
---

```{r setup, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
require(knitr)
require(here)
source(here("data", "environment_vars.R"))
knitr::opts_chunk$set(root.dir = here(), error = TRUE)
knitr::opts_knit$set(echo = TRUE, error = TRUE, root.dir = here())
i = 1


custom_hist = function(x, y, xlab = "", ylab = "", main = "", col = 2)
{
  h = hist(1:(2 * length(x)), breaks = length(x), plot = FALSE)
  
  h$breaks = 2 * (0:length(x))
  h$mids = 2 * (0:(length(x) - 1)) + 1
  h$density = y
  h$counts = y
  plot(h, main = main, xlab = xlab, ylab = ylab, col = col, axes = F)
}

n_bins = 19
x = seq(0, 1, length.out = n_bins)

i = 1;

x_med = c(14, 6, 8, 8, 21, 10, 9, 13, 5, 10, 5, 6, 14)
median(x_med)


```



# iClicker quiz: Question `r colorize(i, "blue")` `r i = i + 1`

Which of the following appears Normally-distributed?
\vfill

```{r, echo=FALSE, fig.width=4, fig.height=1.2}
{
par(mfrow = c(1, 3), mar = c(0, 0, 2, 0))
custom_hist(x, dbeta(x, 7, 2), main = "A")
custom_hist(x, dnorm(x, 0.5, 0.1), main = "B")
custom_hist(x, dbeta(x, 2, 7), main = "C")
}
```

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}





# iClicker quiz: Question  `r colorize(i, "blue")` `r i = i + 1`

What is the **median** value of the following sequence of numbers?
\begin{center}
`r bold_col(sort(x_med), "black")`
\end{center}

\begin{enumerate}[A]
\item 11
\item 9.5
\item 8
\item 9
\item 10
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}



# iClicker quiz: Question `r colorize(i, "blue")` `r i = i + 1`

What is the **3rd Quartile**, also known as the **75th percentile**, rounded to the nearest integer, of the following sequence of numbers?

\begin{center}
`r bold_col(paste0(0:12), "black")`
\end{center}

```{r, include = FALSE}
summary(0:12)
quantile(0:12, 0.75)
```

\begin{enumerate}[A]
\item 6
\item 7
\item 9
\item 11
\item 12
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}





# Announcements

## Salamander groups
- `r colorize("Some students were not included in salamander description groups.", "red")`
- `r colorize("Please verify your group membership and check the Moodle gradebook.", "red")`

## Bear peer-feedback forms
- Extended until tomorrow night.
- Future peer-feedback form deadlines will be enforced.



# Announcements

- This is a short week.
- Chapter 4 has a lot of important information. 
- We'll continue chapter 4 materials into next week.
- The pre-class exercises for next week will include reading questions from chapters 4 and 5.
- Short lecture today
    - I want to have time for the in-class group activity.




# Questions from the salamander exercise

- What is SVL?
- Why is there a `$` in `mander`?
- How is *central tendency* related to the *spread*?
- How do we define *quartiles*?
- Loading data files into R
- In-class R instruction




# Sample Statistics

What are two ways to summarize a collection of numbers?

- Central tendency
- Dispersion
\vfill

Why do we call these *statistics*?
\vfill



# Distributions

What are key features of the **Normal Distribution**?
\vfill

What is the **Uniform Distribution**?
\vfill




# Group Activity:  Random numbers

Self-select groups of 3 or 4.
\vfill

Follow the instructions on Moodle.
\vfill

Submit a single report for the group.
\vfill





<!-- # Central Tendencies -->

<!-- `r bold_col("Mean", "grn")` is the arithmetic average. -->

<!-- * The sum of all the quantities divided by the count of quantities -->

<!-- `r bold_col("Median", "grn")` is the middle value.  There are the same number of observations that are less than and greater than the **`r colorize("median", "grn")`**. -->

<!-- * If there are an *even* number of observations, the median is the average of the 2 central values. -->

<!-- `r bold_col("Mode", "grn")` is the quantity that appears most frequently. -->

<!-- * The mode is not always well-defined, for example if there are no repeated values, or if there are ties. -->



<!-- # Mean -->

<!-- Mean values are often written as variables with a *bar* symbol.  The value of x-bar can be calculated from a collection of quantities **x**. -->

<!-- $$\bar{x} = \frac{\sum{x}_{i}}{n}$$ -->

<!-- How do we read this equantion? -->




<!-- # Properties of the mean -->

<!-- `r colorize("Mean", "grn")` is a good measure of central tendency when data have a **symmetrical** distribution. -->

<!-- * For example, Normally-distributed data -->

<!-- It's easy to calculate the mean value of a vector in R using the funciton `mean()`. -->





<!-- # Median -->

<!-- `r colorize("Median", "grn")` is a good choice when data are skewed. -->

<!-- Why? -->

<!-- In R it's super easy with the `median()` function. -->





<!-- # Mode -->

<!-- `r colorize("Mode", "grn")` is the most commonly occurring quantity. -->

<!-- Mode can be more difficult to use. -->

<!-- What is the mode of the sequence of numbers? -->

<!-- 3, 4, 8, 123, 6, 1239888 -->


<!-- There isn't a simple function in R!  We can write a custom function... but we probably won't need to calculate modes in this course. -->




<!-- # Dispersion -->

<!-- Dispersion, or spread, quantifies the degree to which the individual values in a set of quantities are different from one another. -->

<!-- There are several ways we could measure dispersion.  We've talked about some dispersion measures in class: -->

<!-- * range -->
<!-- * standard deviation (and variance) -->

<!-- Calculating the variance and standard deviation is more complicated than calculating means. -->



<!-- # Variance and standard deviation -->

<!-- Standard deviation is the square root of **`r colorize("Variance", "grn")`. -->

<!-- Let's look at the formula for variance and see if we can figure out why the square root is useful: -->


<!-- $$var(x) = \frac{1}{N}\sum\limits_{i = 1}^N {\left( {x_i - \bar x} \right)^2 } $$ -->

<!-- That looks a lot more complicated than the mean formula! -->

<!-- What are the units of *variance*? -->




<!-- # Variance and standard deviation -->

<!-- If we are measuring height in meters, what are the units of *variance*? -->

<!-- If we measure mass in grams, how can we interpret grams squared? -->

<!-- Variance is difficult to interpret directly.  The square root, i.e. standard deviation, is usually more meaningful. -->




<!-- # Properties of variance and standard deviation -->


<!-- Variance cannot be negative. -->

<!-- A verbal definition of sample variance is:  "The average squared deviation from the mean value." -->

<!-- The *squared deviation* is a key property.  Can the square of a number ever be *negative*? -->

<!-- This definition for variance will make more intuitive sense when we look at Simple Linear Regression. -->




<!-- # Properties of variance and standard deviation -->

<!-- If data are Normally-distributed, the standard deviation has some very useful properties: -->

<!-- * approximately 68% of all observations fall within 1 standard deviation of the mean. -->
<!-- * approximately 95% of all observations fall within 2 standard deviations of the mean. -->










<!-- # Quantiles and Quartiles -->

<!-- Sometimes called *percentiles*. -->

<!-- Quantilies are calculated by placing all of the observations in ascending order. -->

<!-- Let $$x$$ be the value for which we want to calculate a quantile. -->

<!-- The quantile of %%x%% is the count of observations less than %%x%% divided by the total count of observations (*n*). -->

<!-- * Quantiles may be calculated with *n* or *n - 1*. -->




<!-- # Quantiles and quartiles -->

<!-- What is another name for the 50% quantile? -->

<!-- What are quartiles? -->





