---
title: "Week 7: Tests for Differences"
subtitle:  "Session 1"
date: Spring 2020
output:
  beamer_presentation:
    pandoc_args: !expr paste0(here::here('css', 'beamer.yaml'))
header-includes:
  \input{`r here::here("css", "headers_tikz.tex")`}
---

```{r setup, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
require(knitr)
require(here)
source(here("data", "environment_vars.R"))
knitr::opts_chunk$set(root.dir = here(), error = TRUE)
knitr::opts_knit$set(echo = TRUE, error = TRUE, root.dir = here())
i = 1
```


## iClicker Question `r colorize(i, "blue")` `r i = i + 1`

Scenario: 

- You want to know whether the growth rates of oak trees differ in two habitats.  
- You have two clones each of 37 different genotypes.  
- One clone of each genotype is planted in a mesic prairie, the other in an old agricultural field.

- Which test should you use?

\begin{enumerate}[A]
\item \texttt{two-sample t-test}
\item \texttt{Mann-Whitney paried test}
\item \texttt{Mann-Whitney U test}
\item \texttt{paired t-test}
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}



## iClicker Question `r colorize(i, "blue")` `r i = i + 1`

Scenario: 

- You are testing whether Pennsylvania sedge (C. pennsylvanica) cover is greater under hardwoods or conifers. 
- You have collected percent cover data from 50 hardwood and 50 conifer patches at Mount Toby.

- Which test should you use?

\begin{enumerate}[A]
\item \texttt{two-sample t-test}
\item \texttt{Mann-Whitney paried test}
\item \texttt{Mann-Whitney U test}
\item \texttt{paired t-test}
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}



## iClicker Question `r colorize(i, "blue")` `r i = i + 1`

Scenario: 

-You want to know whether the growth rates of oak trees differ in two habitats.  

-You think they will grow more quickly in a old agricultural field than in a mesic prairie.

- Which is the best alternative hypothesis?

\begin{enumerate}[A]
\item \texttt{growth is faster in the prairie}
\item \texttt{there is no difference in growth between habitats}
\item \texttt{growth is faster in the field}
\item \texttt{there is a difference in growth between habitats}
\item \texttt{growth rates are the same}
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}



## iClicker Question `r colorize(i, "blue")` `r i = i + 1`

Scenario: 

-You want to know whether the growth rates of oak trees differ in two habitats: an old agricultural field and mesic prairie.

- Which is the best alternative hypothesis?

\begin{enumerate}[A]
\item \texttt{growth is faster in the prairie}
\item \texttt{there is no difference in growth between habitats}
\item \texttt{growth is faster in the field}
\item \texttt{there is a difference in growth between habitats}
\item \texttt{growth rates are the same}
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}



## iClicker Question `r colorize(i, "blue")` `r i = i + 1`

Scenario: 

- You calculate a t-value, and an associated p-value, for individual length in two populations of Daphnia.  
- Which critical significance category for the t-value would give you the best evidence that the two populations are different:

\begin{enumerate}[A]
\item \texttt{1\%}
\item \texttt{10\%}
\item \texttt{5\%}
\item \texttt{95\%}
\item \texttt{0.1\%}
\end{enumerate}

\logoSoutheast{`r here::here("slides", "slide_images", "iClicker_logo.png")`}






## If you're feeling stuck, remember my office hours are Tuesday/Thursday 1:00 - 2:00.

\begin{center}
\includegraphics[width = 0.95\textwidth]{`r here("slides", "slide_images", "betsy_tangled.jpg")`}
\end{center}




## For Today

- Toward statistics

- Tests for differences

### Follow-up questions from the Chapter 6 homework

- What questions do you have?




## Beyond graphs, Towards statistics

- Graphs are powerful tools that provide insight and understanding of the patterns and relationships in the data.

- Graphs alone don't give us the complete answer.  We need to **quantify** the relationships we see in our plots.

- What other tools do we have to **support** our conclusions?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift = -2cm] at (current page)
{\includegraphics[height=2in]{`r here("slides", "slide_images", "QM.png")`}};
\end{tikzpicture}


## Beyond graphs, Towards statistics

- How can we **quantify** our evidence for relationships?
    - Are differences between groups *significant*?
    - Are differences between groups *meaningful*?
    
\begin{tikzpicture}[remember picture,overlay]
\node[yshift = 4cm] at (current page.south)
{\includegraphics[height = 7cm]{`r here("slides", "slide_images", "salbox.pdf")`}};
\end{tikzpicture}




## Beyond graphs, Towards statistics

- How can we **quantify** our evidence for relationships?
    - Are associations between 2 variables *significatnt*?
    - Are associations between 2 variables *meaningful*?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift = 4cm] at (current page.south)
{\includegraphics[height = 7cm]{`r here("slides", "slide_images", "salscatter.pdf")`}};
\end{tikzpicture}




## Beyond graphs, Towards statistics
   
- Statistics is the tool we use to formally answer these questions!
    - Are differences *are*/*are not* significant?
    - Are associations *are*/*are not* significant?
    
### Wait a second... what do we mean when we say **significant**?
    
\begin{tikzpicture}[remember picture,overlay]
\node[yshift = -2cm] at (current page)
{\includegraphics[height=2in]{`r here("slides", "slide_images", "QM.png")`}};
\end{tikzpicture}


## Let's examine some plots to gain intuition:

- Scenario:  We want to know whether the size of 3-year-old bluegill (*Lepomis macrochirus*) are larger in some Massachusetts lakes than others. 

- We have collected data for bluegill from Wyola Lake and the Quabbin Reservoir in Western Mass.


\begin{tikzpicture}[remember picture,overlay]
\node[yshift = 2.9cm] at (current page.south)
{\includegraphics[height = 4.5cm]{`r here("slides", "slide_images", "bluegill.jpg")`}};
\end{tikzpicture}

\footnotetext[1]{Image credit: New York Fish and Game Commission}



## Bluegill Data I

- Are differences between lakes *significant*?
- Are differences between lakes *meaningful*?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift = 3.4cm] at (current page.south)
{\includegraphics[height = 6.5cm]{`r here("slides", "slide_images", "quab_wyola_1.pdf")`}};
\end{tikzpicture}



## Bluegill Data II

- Are differences between lakes *significant*?
- Are differences between lakes *meaningful*?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift = 3.4cm] at (current page.south)
{\includegraphics[height = 6.5cm]{`r here("slides", "slide_images", "quab_wyola_2.pdf")`}};
\end{tikzpicture}



## Bluegill Data III

- Are differences between lakes *significant*?
- Are differences between lakes *meaningful*?
\begin{tikzpicture}[remember picture,overlay]
\node[yshift = 3.4cm] at (current page.south)
{\includegraphics[height = 6.5cm]{`r here("slides", "slide_images", "quab_wyola_3.pdf")`}};
\end{tikzpicture}



## Bluegill Data IV

- Are differences between lakes *significant*?
- Are differences between lakes *meaningful*?
\begin{tikzpicture}[remember picture,overlay]
\node[yshift = 3.4cm] at (current page.south)
{\includegraphics[height = 6.5cm]{`r here("slides", "slide_images", "quab_wyola_4.pdf")`}};
\end{tikzpicture}










## Tests for differences

- How can we test if two groups of observations are different?

- What kinds of tests do we know about?

- Which statistic(s) do the tests test for?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1.5cm] at (current page)
{\includegraphics[height=2in]{`r here("slides", "slide_images", "QM.png")`}};
\end{tikzpicture}


## Tests for differences

Often we want to know if two of more samples are different

- are the sample *means* different?
- are the sample *medians* different?
- are the differences *statistically significant*?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1.8cm] at (current page)
{\includegraphics[height=2in]{`r here("slides", "slide_images", "meansA.pdf")`}};
\end{tikzpicture}

## Tests for differences

Often we want to know if two of more samples are different

- are the sample *means* different?
- are the sample *medians* different?
- are the differences *statistically significant*?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1.5cm] at (current page)
{\includegraphics[height=2in]{`r here("slides", "slide_images", "means2.pdf")`}};
\end{tikzpicture}

## Tests for differences

Often we want to know if two of more samples are different

- are the sample *means* different?
- are the sample *medians* different?
- are the differences *statistically significant*?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1.5cm] at (current page)
{\includegraphics[height=2in]{`r here("slides", "slide_images", "means3.pdf")`}};
\end{tikzpicture}

## Tests for differences

Often we want to know if two of more samples are different

- are the sample *means* different?
- are the sample *medians* different?
- are the differences *statistically significant*?

To determine the significance of differences between **two**, we need a statistical test

- *t-test*
- *U-test*


## Tests for differences: intuition

- What information would we need to know?
- What kinds of evidence would support our conclusion?
- How do we define *different*?

\vfill

- Let's draw some distributions:



## Tests for differences: intuition with R

```{r, eval = FALSE, echo = FALSE, include = FALSE}
setwd(here())
ls()
dir.exists("data")
file.exists("data/mander.csv")
dat = read.csv("data/mander.csv")
head(dat)

subset(dat, Sex == "female")
boxplot(dat$SVL ~ dat$Sex)

t.test(dat$SVL ~ dat$Sex)

```







## Differences: t-test

Purpose:

- compare the means of two samples (say $a$ and $b$)

Assumptions:

- both samples normally distributed
- both samples have equal variances

## Differences: t-test

Purpose:

- compare the means of two samples (say $a$ and $b$)

Assumptions:

- both samples normally distributed
- both samples have equal variances

$$t = \frac{|\bar{x}_a - \bar{x}_b|}{\sqrt{\frac{s^2_a}{n_a}+\frac{s^2_b}{n_b}}} $$

- $t$: the *t*-statistic
- $\bar{x}$: sample mean
- $s$: sample standard deviation
- $n$: sample size

## Differences: t-test

Purpose:

- compare the means of two samples (say $a$ and $b$)

Assumptions:

- both samples normally distributed
- both samples have equal variances

$$t = \frac{|\bar{x}_a - \bar{x}_b|}{\sqrt{\frac{s^2_a}{n_a}+\frac{s^2_b}{n_b}}} $$

- if $|\bar{x}_a - \bar{x}_b|$ is large, then $t$ is \textcolor{blue}{????}

\vspace{0.2cm}

- if $\sqrt{\frac{s^2_a}{n_a}+\frac{s^2_b}{n_b}}$ is large, then $t$ is \textcolor{blue}{????}

## Differences: t-test

Purpose:

- compare the means of two samples (say $a$ and $b$)

Assumptions:

- both samples normally distributed
- both samples have equal variances

$$t = \frac{|\bar{x}_a - \bar{x}_b|}{\sqrt{\frac{s^2_a}{n_a}+\frac{s^2_b}{n_b}}} $$

- if $|\bar{x}_a - \bar{x}_b|$ is large, then $t$ is \textcolor{blue}{large}

\vspace{0.2cm}

- if $\sqrt{\frac{s^2_a}{n_a}+\frac{s^2_b}{n_b}}$ is large, then $t$ is \textcolor{blue}{small}


## Differences: t-test

\normalsize

Understanding the *t-distribution*:

\begin{tikzpicture}[remember picture, overlay]
\node[yshift = -1cm] at (current page)
{\includegraphics[height = 2in]{`r here("slides", "slide_images", "t5.pdf")`}};
\end{tikzpicture}

## Differences: t-test

Understanding the *t-distribution*:

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1cm] at (current page)
{\includegraphics[height=2in]{`r here("slides", "slide_images", "t25.pdf")`}};
\end{tikzpicture}

## Differences: t-test

Understanding the *t-distribution*:

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1cm] at (current page)
{\includegraphics[height=2in]{`r here("slides", "slide_images", "t50.pdf")`}};
\end{tikzpicture}

## Differences: t-test

Understanding the *t-distribution*:

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1cm] at (current page)
{\includegraphics[height=2in]{`r here("slides", "slide_images", "samplesize.pdf")`}};
\end{tikzpicture}

## Differences: t-test

Understanding the *t-distribution*:

- whether a difference is significant depends on:
    - the *t-statistic*
    - degrees-of-freedom ($n_a - 1 + n_b -1$)

- larger *t-statistics* more likely to be significant

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2.4cm, yshift=-2.3cm] at (current page)
{\includegraphics[height=2in]{`r here("slides", "slide_images", "t50.pdf")`}};
\end{tikzpicture}

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=2.4cm, yshift=-2.3cm] at (current page)
{\includegraphics[height=2in]{`r here("slides", "slide_images", "samplesize.pdf")`}};
\end{tikzpicture}

## Differences: t-test

Understanding the *p-value*:

- *p-value* is the probability of observing a *t-statistic* as high as we did by chance
- if *p-value* is lower than significance level (e.g. 5%):
    - difference is significant
    - reject the null hypothesis
    - accept the alternative hypothesis

## Differences: t-test

Which *t-test*?

- standard *t-test*
    - compare two independent samples
    - both normally distributed
    - equal (similar) variances
    - samples sizes can be the same or not

$$t = \frac{|\bar{x}_a - \bar{x}_b|}{\sqrt{\frac{s^2_a}{n_a}+\frac{s^2_b}{n_b}}} $$

- $t$: the *t*-statistic
- $\bar{x}$: sample mean
- $s$: sample standard deviation
- $n$: sample size

## Differences: paired t-test

Sometimes samples are not independent

- compare pairs of samples
    - e.g., before-after
    - e.g., north-south
    - e.g., left-right
- both normally distributed
- equal (similar) variances
- samples sizes *must* be the

## Differences: paired t-test

Which *t-test*?

- paired *t-test*
    - compare pairs of samples
    - both normally distributed
    - equal (similar) variances
    - samples sizes are ______ ?

$$t = \frac{\bar{D}}{\sqrt{\frac{s^2_D}{n}}} $$

- $t$: the *t*-statistic
- $\bar{D}$: mean of the *differences*
- $s$: standard deviation of the *differences*
- $n$: number of *paired* samples

## Differences: When might the t-test be inappropriate?

\begin{tikzpicture}[remember picture,overlay]
\node at (current page)
{\includegraphics[height=2in]{`r here("slides", "slide_images", "QM.png")`}};
\end{tikzpicture}

## Differences: U-test

- compare two samples
- one or both *not* normally distributed
- based on *median*, *range*, and *ranks*
- rank all values as one sample, calculate group rank sums $R$
- calculate a *U*-value, a measure of overlap

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=3cm] at (current page.south)
{\includegraphics[height=1.8in]{`r here("slides", "slide_images", "distributions.jpg")`}};
\end{tikzpicture}



## Differences: U-test

- compare two samples
- both or differences *not* normally distributed
- based on *median*, *range*, and *ranks*
- rank all values as one sample, calculate group rank sums $R$
- calculate a *U*-value, a measure of overlap

$$U_a = n_a \times n_b + \frac{n_a(n_a+1)}{2}-R_a $$
$$U_b = n_b \times n_a + \frac{n_b(n_b+1)}{2}-R_b $$

- $n_a$: number of samples in sample $a$
- $n_b$: number of samples in sample $b$
- $R_a$: sum of the ranks of values in $a$
- $R_b$: sum of the ranks of values in $b$

## Differences: U-test

- compare two samples
- both or differences *not* normally distributed
- based on *median*, *range*, and *ranks*
- rank all values as one sample, calculate group rank sums $R$
- calculate a *U*-value, a measure of overlap

$$U_a = n_a \times n_b + \frac{n_a(n_a+1)}{2}-R_a $$
$$U_b = n_b \times n_a + \frac{n_b(n_b+1)}{2}-R_b $$

- smallest is used to find the $p$-value
- unlike the t-statistic, lower U-values are more likely to be significant

## Differences: Wilcoxon matched-pairs test

- both or differences *not* normally distributed
- based on ranked *differences*
    - first calculate the differences
    - second rank the differences
    - 0's not ranked
- sum and compare +ve and -ve ranks

$$W^+ = \sum R^+$$
$$W^- = \sum R^-$$

- $W^+$: the Wilcoxon test statistic for positive differences
- $W^+$: the Wilcoxon test statistic for negative differences
- $R^+$: the sum of the ranks of positive differences
- $R^+$: the sum of the ranks of negative differences

## Differences: Wilcoxon matched-pairs test

- pairs or differences *not* normally distributed
- based on ranked *differences*
    - first calculate the differences
    - second rank the differences
    - 0's not ranked
- sum and compare +ve and -ve ranks

$$W^+ = \sum R^+$$
$$W^- = \sum R^-$$

- smallest is used to find the $p$-value
- lower *W*-values are more likely to be significant

## Group Assignment

Using the whale count data, complete the group assignment on Moodle.