---
title: "Tests for Differences"
author: |
  | Introduction to Quantitative Ecology
  | Fall 2016
  | Chris Sutherland
  | csutherland@umass.edu

header-includes:
   - \usepackage{tabto}
   - \usepackage{verbatim}
   - \usepackage{amsmath}
   - \usepackage{mathtools}
   - \usepackage{graphicx}
   - \usepackage{tikz}
   - \usepackage{tikzpagenodes}
   - \definecolor{OG}{RGB}{0,64,8}
   - \definecolor{LG}{RGB}{0,102,51}
   - \definecolor{myRed}{RGB}{228,26,28}
   - \definecolor{myBlue}{RGB}{55,126,184}
   - \definecolor{myGreen}{RGB}{77,175,74}
   - \definecolor{myPurple}{RGB}{152,78,163}
   - \setbeamercolor{itemize item}{fg=white!0!LG}
   - \setbeamercolor{enumerate item}{fg=white!0!LG}
   - \setbeamercolor{enumerate subitem}{fg=white!70!LG}
   - \setbeamercolor{itemize subitem}{fg=white!70!LG}   
   - \setbeamercolor{itemize subsubitem}{fg=white!70!LG}   
   - \setbeamercolor{navigation symbols}{fg=white!70!LG, bg=white!70!LG}
   - \usepackage{inputenc}
   - \usepackage{booktabs}
   - \usepackage{caption}
   - \usetikzlibrary{patterns,arrows,decorations.pathreplacing}
   - \usepackage{setspace}
   - \DeclareGraphicsExtensions{.pdf,.png,.jpg,.gif}
   - \usepackage{fixmath}
   - \usepackage{tabto}
   - \graphicspath{{"/Users/megmaclean/Box/Courses/NRC_290b_2019_Fall/Images/Diffs/"}}

output:
  beamer_presentation:
    highlight: tango
    theme: "default"
    colortheme: "spruce"
    fonttheme: "serif"
    slide_level: 2
    incremental: false
classoption: t
---
## Group evaluations

\vfill

1. *t-test*s are used to test for differences between what?

2. When do you need to use the *U-test*?

3. If sample $a$ has $n_a = 10$ samples, and sample $b$ has $n_b=10$ samples, then what is the degrees of freedom?

4. Conducting a statistical test of differences using a t-test, you get a $p$-value or $p=0.02$. Using a 5% significance level, what would you conclude?

5. Would a large or a small *t-statistic* be more likely to indicate a significant difference between the means of two samples, and why?


```{r setup, echo=FALSE, results='hide', message=FALSE}
library(knitr)
opts_knit$set(root.dir = "/Users/megmaclean/Box/Courses/NRC_290b_2019_Fall/")
#opts_knit$set(root.dir = "C:/Users/csutherland/Dropbox/C/Courses class/NRC290b/NRC_290b_2018/")
options(width=120)
options(width.cutoff=80)
opts_chunk$set(prompt = FALSE, comment = NA, mysize=TRUE, size="\\scriptsize")
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knit_hooks$set(document = hook2)
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
```

```{r, echo=F, warning=FALSE, results='hide', message=FALSE}
classData <- read.table(file="data/classData.txt")
mander <- read.table(file="data/mander.txt",h=T)
bp <- read.table(file="data/bearpoop.txt",h=T)
```


```{r,eval=TRUE,echo=FALSE,results="hide"}
x1 <- rnorm(100000,50,10)
x2 <- rnorm(100000,150,10)
x3 <- rnorm(100000,52,10)
x4 <- rnorm(100000,60,10)

pdf("Images/Diffs/meansA.pdf",height=4,width=5)
par(mfrow=c(2,1),oma=c(0,0,0,0),mar=c(2,2,1,1))
plot(density(x1),col=1,lwd=1,xlim=c(0,200),ylim=c(0,0.05),
     main="",axes=F,xlab="",ylab="")
abline(v=mean(x1),lwd=2,col=2)
abline(v=mean(x2),lwd=2,col=4)
lines(density(x1),col=1, lwd=1)
lines(density(x2),col=1, lwd=1)
axis(1)
box(bty="l")
boxplot(cbind(x1,x2),ylim=c(0,200),horizontal=TRUE,pch="",
        col=adjustcolor(c(2,4),0.3),axes=F)
axis(1)
abline(v=mean(x1),lwd=2,col=2)
abline(v=mean(x2),lwd=2,col=4)
boxplot(cbind(x1,x2),ylim=c(0,200),horizontal=TRUE,pch="",
        col=adjustcolor(c(2,4),0.3),axes=F,add=T)
box(bty="l")
dev.off()

pdf("Images/Diffs/means2.pdf",height=4,width=5)
par(mfrow=c(2,1),oma=c(0,0,0,0),mar=c(2,2,1,1))
plot(density(x1),col=1,lwd=1,xlim=c(0,200),ylim=c(0,0.05),
     main="",axes=F,xlab="",ylab="")
abline(v=mean(x1),lwd=2,col=2)
abline(v=mean(x3),lwd=2,col=4)
lines(density(x1),col=1, lwd=1)
lines(density(x3),col=1, lwd=1)
axis(1)
box(bty="l")
boxplot(cbind(x1,x3),ylim=c(0,200),horizontal=TRUE,pch="",
        col=adjustcolor(c(2,4),0.3),axes=F)
axis(1)
abline(v=mean(x1),lwd=2,col=2)
abline(v=mean(x3),lwd=2,col=4)
boxplot(cbind(x1,x3),ylim=c(0,200),horizontal=TRUE,pch="",
        col=adjustcolor(c(2,4),0.3),axes=F,add=T)
box(bty="l")
dev.off()

pdf("Images/Diffs/means3.pdf",height=4,width=5)
par(mfrow=c(2,1),oma=c(0,0,0,0),mar=c(2,2,1,1))
plot(density(x1),col=1,lwd=1,xlim=c(0,200),ylim=c(0,0.05),
     main="",axes=F,xlab="",ylab="")
abline(v=mean(x1),lwd=2,col=2)
abline(v=mean(x4),lwd=2,col=4)
lines(density(x1),col=1, lwd=1)
lines(density(x4),col=1, lwd=1)
axis(1)
box(bty="l")
boxplot(cbind(x1,x4),ylim=c(0,200),horizontal=TRUE,pch="",
        col=adjustcolor(c(2,4),0.3),axes=F)
axis(1)
abline(v=mean(x1),lwd=2,col=2)
abline(v=mean(x4),lwd=2,col=4)
boxplot(cbind(x1,x4),ylim=c(0,200),horizontal=TRUE,pch="",
        col=adjustcolor(c(2,4),0.3),axes=F,add=T)
box(bty="l")
dev.off()
```

## Tests for differences

\begin{tikzpicture}[remember picture,overlay]
\node at (current page)
{\includegraphics[height=2in]{QM.png}};
\end{tikzpicture}


## Tests for differences

Often we want to know if two of more samples are different

- are the sample *means* different?
- are the sample *medians* different?
- are the differences *statistically significant*?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1.5cm] at (current page)
{\includegraphics[height=2in]{meansA.pdf}};
\end{tikzpicture}

## Tests for differences

Often we want to know if two of more samples are different

- are the sample *means* different?
- are the sample *medians* different?
- are the differences *statistically significant*?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1.5cm] at (current page)
{\includegraphics[height=2in]{means2.pdf}};
\end{tikzpicture}

## Tests for differences

Often we want to know if two of more samples are different

- are the sample *means* different?
- are the sample *medians* different?
- are the differences *statistically significant*?

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1.5cm] at (current page)
{\includegraphics[height=2in]{means3.pdf}};
\end{tikzpicture}

## Tests for differences

Often we want to know if two of more samples are different

- are the sample *means* different?
- are the sample *medians* different?
- are the differences *statistically significant*?

To determine the significance of differences between **two**, we need a statistical test

- *t-test*
- *U-test*

## Differences: t-test

Purpose:

- compare the means of two samples (say $a$ and $b$)

Assumptions:

- both samples normally distributed
- both samples have equal variances

## Differences: t-test

Purpose:

- compare the means of two samples (say $a$ and $b$)

Assumptions:

- both samples normally distributed
- both samples have equal variances

$$t = \frac{|\bar{x}_a - \bar{x}_b|}{\sqrt{\frac{s^2_a}{n_a}+\frac{s^2_b}{n_b}}} $$

- $t$: the *t*-statistic
- $\bar{x}$: sample mean
- $s$: sample standard deviation
- $n$: sample size

## Differences: t-test

Purpose:

- compare the means of two samples (say $a$ and $b$)

Assumptions:

- both samples normally distributed
- both samples have equal variances

$$t = \frac{|\bar{x}_a - \bar{x}_b|}{\sqrt{\frac{s^2_a}{n_a}+\frac{s^2_b}{n_b}}} $$

- if $|\bar{x}_a - \bar{x}_b|$ is large, then $t$ is \textcolor{blue}{????}

\vspace{0.2cm}

- if $\sqrt{\frac{s^2_a}{n_a}+\frac{s^2_b}{n_b}}$ is large, then $t$ is \textcolor{blue}{????}

## Differences: t-test

Purpose:

- compare the means of two samples (say $a$ and $b$)

Assumptions:

- both samples normally distributed
- both samples have equal variances

$$t = \frac{|\bar{x}_a - \bar{x}_b|}{\sqrt{\frac{s^2_a}{n_a}+\frac{s^2_b}{n_b}}} $$

- if $|\bar{x}_a - \bar{x}_b|$ is large, then $t$ is \textcolor{blue}{large}

\vspace{0.2cm}

- if $\sqrt{\frac{s^2_a}{n_a}+\frac{s^2_b}{n_b}}$ is large, then $t$ is \textcolor{blue}{small}


## Differences: t-test

```{r,eval=TRUE,echo=FALSE,results="hide"}

tt <- seq(-3,3,0.1)

pdf("Images/Diffs/t5.pdf",height=4,width=5)
par(mfrow=c(1,1),oma=c(0,0,0,0),mar=c(4,4,1,1))
plot(tt,dt(tt,df=1),type="l",lwd=2,col=4,xlab="",ylim=c(0,0.4),ylab="Probability",main="Degrees of Freedom = 1", las=1,bty="l")
abline(h=0.05,lty=2,col=2)
dev.off()

pdf("Images/Diffs/t25.pdf",height=4,width=5)
par(mfrow=c(1,1),oma=c(0,0,0,0),mar=c(4,4,1,1))
plot(tt,dt(tt,df=1),type="l",col=1,lwd=1,xlab="",ylim=c(0,0.4),ylab="Probability",main="Degrees of Freedom = 5", las=1,bty="l")
lines(tt,dt(tt,df=5),type="l",lwd=2,col=4,xlab="",ylab="")
abline(h=0.05,lty=2,col=2)
dev.off()

pdf("Images/Diffs/t50.pdf",height=4,width=5)
par(mfrow=c(1,1),oma=c(0,0,0,0),mar=c(4,4,1,1))
plot(tt,dt(tt,df=1),type="l",col=1,lwd=1,xlab="",ylim=c(0,0.4),ylab="Probability",main="Degrees of Freedom = 500", las=1,bty="l")
lines(tt,dt(tt,df=5),type="l",col=1,lwd=1,xlab="",ylab="")
lines(tt,dt(tt,df=500),type="l",lwd=2,col=4,xlab="",ylab="")
abline(h=0.05,lty=2,col=2)
dev.off()

df <- 1:50
pdf("Images/Diffs/samplesize.pdf",height=4,width=5)
par(mfrow=c(1,1),oma=c(0,0,0,0),mar=c(4,4,1,1))
plot(df,qt(0.975, df=df), type="l", lwd=2, xlab="Degrees-of-Freedom",
     ylab="Critical t value", las=1, bty="l", ylim=c(0,13))
points(df,qt(0.975, df=df), pch=16, col=4)
dev.off()
```

\normalsize

Understanding the *t-distribution*:

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1cm] at (current page)
{\includegraphics[height=2in]{t5.pdf}};
\end{tikzpicture}

## Differences: t-test

Understanding the *t-distribution*:

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1cm] at (current page)
{\includegraphics[height=2in]{t25.pdf}};
\end{tikzpicture}

## Differences: t-test

Understanding the *t-distribution*:

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1cm] at (current page)
{\includegraphics[height=2in]{t50.pdf}};
\end{tikzpicture}

## Differences: t-test

Understanding the *t-distribution*:

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=-1cm] at (current page)
{\includegraphics[height=2in]{samplesize.pdf}};
\end{tikzpicture}

## Differences: t-test

Understanding the *t-distribution*:

- whether a difference is significant depends on:
    - the *t-statistic*
    - degrees-of-freedom ($n_a - 1 + n_b -1$)
    
- larger *t-statistics* more likely to be significant

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2.4cm, yshift=-2.3cm] at (current page)
{\includegraphics[height=1.5in]{t50.pdf}};
\end{tikzpicture}

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=2.4cm, yshift=-2.3cm] at (current page)
{\includegraphics[height=1.5in]{samplesize.pdf}};
\end{tikzpicture}

## Differences: t-test

Understanding the *p-value*:

- *p-value* is the probability of observing a *t-statistic* as high as we did by chance
- if *p-value* is lower than significance level (e.g. 5%):
    - difference is significant
    - reject the null hypothesis
    - accept the alternative hypothesis

## Differences: t-test

Which *t-test*?

- standard *t-test*
    - compare two independent samples
    - both normally distributed
    - equal (similar) variances
    - samples sizes can be the same or not
    
$$t = \frac{|\bar{x}_a - \bar{x}_b|}{\sqrt{\frac{s^2_a}{n_a}+\frac{s^2_b}{n_b}}} $$

- $t$: the *t*-statistic
- $\bar{x}$: sample mean
- $s$: sample standard deviation
- $n$: sample size

## Differences: paired t-test

Sometimes samples are not independent

- compare pairs of samples
    - e.g., before-after
    - e.g., north-south
    - e.g., left-right
- both normally distributed
- equal (similar) variances
- samples sizes *must* be the

## Differences: paired t-test

Which *t-test*?

- paired *t-test*
    - compare pairs of samples
    - both normally distributed
    - equal (similar) variances
    - samples sizes *must* be the
    
$$t = \frac{\bar{D}}{\sqrt{\frac{s^2_D}{n}}} $$

- $t$: the *t*-statistic
- $\bar{D}$: mean of the *differences*
- $s$: standard deviation of the *differences*
- $n$: number of *paired* samples

## Differences: U-test

\begin{tikzpicture}[remember picture,overlay]
\node at (current page)
{\includegraphics[height=2in]{QM.png}};
\end{tikzpicture}

## Differences: U-test

- compare two samples
- both *not* normally distributed
- based on *median*, *range*, and *ranks*
- rank all values as one sample, calculate group rank sums $R$
- calculate a *U*-value, a measure of overlap

\begin{tikzpicture}[remember picture,overlay]
\node[yshift=3cm] at (current page.south)
{\includegraphics[height=1.8in]{distributions.jpg}};
\end{tikzpicture}

    
## Differences: U-test

- compare two samples
- both or differences *not* normally distributed
- based on *median*, *range*, and *ranks*
- rank all values as one sample, calculate group rank sums $R$
- calculate a *U*-value, a measure of overlap

$$U_a = n_a \times n_b + \frac{n_a(n_a+1)}{2}-R_a $$
$$U_b = n_b \times n_a + \frac{n_b(n_b+1)}{2}-R_b $$

- $n_a$: number of samples in sample $a$
- $n_b$: number of samples in sample $b$
- $R_a$: sum of the ranks of values in $a$
- $R_b$: sum of the ranks of values in $b$

## Differences: U-test

- compare two samples
- both or differences *not* normally distributed
- based on *median*, *range*, and *ranks*
- rank all values as one sample, calculate group rank sums $R$
- calculate a *U*-value, a measure of overlap

$$U_a = n_a \times n_b + \frac{n_a(n_a+1)}{2}-R_a $$
$$U_b = n_b \times n_a + \frac{n_b(n_b+1)}{2}-R_b $$

- smallest is used to find the $p$-value
- unlike the t-statistic, lower U-values are more likely to be significant

## Differences: Wilcoxon matched-pairs test

- both or differences *not* normally distributed
- based on ranked *differences*
    - first calculate the differences
    - second rank the differences
    - 0's not ranked
- sum and compare +ve and -ve ranks
    
$$W^+ = \sum R^+$$
$$W^- = \sum R^-$$

- $W^+$: the Wilcoxon test statistic for positive differences
- $W^+$: the Wilcoxon test statistic for negative differences
- $R^+$: the sum of the ranks of positive differences
- $R^+$: the sum of the ranks of negative differences

## Differences: Wilcoxon matched-pairs test

- pairs or differences *not* normally distributed
- based on ranked *differences*
    - first calculate the differences
    - second rank the differences
    - 0's not ranked
- sum and compare +ve and -ve ranks
    
$$W^+ = \sum R^+$$
$$W^- = \sum R^-$$

- smallest is used to find the $p$-value
- lower *W*-values are more likely to be significant
   
## Group Assignment

Using the whale count data, compare the differences between first and second abundance guesstimates using first `excel` and then `R`.

Submit a single written group report that outlines the following points:

1. state the null and alternative hypotheses being tested
2. the reason for choosing the statistical test you used 
3. a summary of the results:
    - degrees-of-freedom, test statistic, p-values (at 5% level)?
    - did you accept or reject the null hypothesis?
    - is there a difference?
4. conduct the analysis in `R` and `excel` and submit:
    - a written report of points 1, 2 and 3 as **PDF**
    - an `excel` workbook showing your results
    - a saved `R` file showing results
