---
title: "Comparing whale guesstimates"
author: "Chris Sutherland"
header-includes:
  - \usepackage{hyperref}
  - \usepackage[onehalfspacing]{setspace}
  - \usepackage{graphicx}
  - \DeclareGraphicsExtensions{.pdf,.png,.jpg}
  - \usepackage{comment}
  - \usepackage[x11names]{xcolor}
  - \usepackage{sectsty}
  - \definecolor{brn}{rgb}{.6,0,0}
  - \sectionfont{\color{brn}}
  - \subsectionfont{\color{brn}\slshape}
  - \DeclareUnicodeCharacter{00A0}{ }
  - \usepackage{booktabs}
  - \usepackage{multicol}
  - \usepackage{tikz}
  - \usepackage{tabto}
output:
  pdf_document:
    fig_caption: yes
    keep_tex: no
    number_sections: no
    toc: no
documentclass: article
fontsize: 12pt
font: serif
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Group Assignment questions:

Using the whale count data, compare the differences between first and second abundance guesstimates using R and excel and report the the following

1. state the null and alternative hypotheses being tested
2. the reason for choosing the statistical test you used
3. a summary of the results:
    - degrees-of-freedom, test statistic, p-values (at 5% level)?
    - did you accept or reject the null hypothesis?
    - is there a difference?
4. conduct the analysis in `R` and `excel` and submit:
    - a written report of points 1, 2 and 3 as PDF
    - an excel workbook showing your results
    - a saved R file showing results

## Getting started

First we need to read in the data. While we are at it, let's look at the first 6 rows using the `head()` function:

```{r}
whales <- read.csv("whales.csv")
head(whales)
```

# Which test and why?

Here the objective is to compare two samples. These samples actually represent *pairs* of values: each person provided an initial and an updated guesstimate of whale abundance. So, this means we proceed as follows:

1. Hypotheses:

    - the null hypothesis is that there is *no difference* between the initial and updated guesstimates
    - the alternative hypothesis is that there *is a difference* between the initial and updated guesstimates
    - we can test these hypotheses using paired t-test

2. Why the paired *t*-test?

    - data are pirs of values from the same individual, i.e., they are not independent
    - data are normally distributed:

Can look at a histogram - looking for a symmetrical distribution:

```{r, echo=TRUE, eval=FALSE}
par(mfrow=c(1,3),oma=c(0,0,0,0),mar=c(4,4,2,2))
hist(whales$estimate)                  # first sample
hist(whales$estimate2)                 # second sample
hist(whales$estimate-whales$estimate2) # differences
```

```{r, echo=FALSE, eval=TRUE, fig.align='center',fig.height=2.4,fig.width=7.2}
par(mfrow=c(1,3),oma=c(0,0,0,0),mar=c(4,4,2,2))
hist(whales$estimate, col=adjustcolor(2,0.3), main="Guesstimate 1")
hist(whales$estimate2, col=adjustcolor(3,0.3), main="Guesstimate 2")
hist(whales$estimate-whales$estimate2, col=adjustcolor(4,0.3), main="Difference (1-2)")
```

Can formally evaluate using the `shapiro.test()` - looking for a *p*-value > 0.05, i.e., not significantly different from normally distributed: 
    
```{r}
shapiro.test(whales$estimate)                 # first sample
shapiro.test(whales$estimate2)                # second sample
shapiro.test(whales$estimate-whales$estimate2) # differences
```



my.t.test <- t.test(whales$estimate,whales$estimate2,paired=TRUE)


```
    
    
    the reason for choosing the statistical test you used
3. a summary of the results:
I degrees-of-freedom, test statistic, p-values (at 5% level)?
I did you accept or reject the null hypothesis?
I is there a difference?
4. conduct the analysis in R and excel and submit:
I a written report of points 1, 2 and 3 as PDF
I an excel workbook showing your results
I a saved R file showing results


# are the samples paired?
# yes!

# are the samples normally distributed?

shapiro.test(whales$estimate) ## NOT SIG DIFF FROM NORMAL!

#data:  whales$estimate
#W = 0.96042, p-value = 0.1731

shapiro.test(whales$estimate2) ## NOT SIG DIFF FROM NORMAL!

#data:  whales$estimate2
#W = 0.9674, p-value = 0.2969

my.t.test <- t.test(whales$estimate,whales$estimate2,paired=TRUE)


Using the whale count data, compare the differences between
first and second abundance guesstimates using R and excel.
Submit a single written group report that outlines the following
points:
1. state the null and alternative hypotheses being tested
2. the reason for choosing the statistical test you used
3. a summary of the results:
I degrees-of-freedom, test statistic, p-values (at 5% level)?
I did you accept or reject the null hypothesis?
I is there a difference?
4. conduct the analysis in R and excel and submit:
I a written report of points 1, 2 and 3 as PDF
I an excel workbook showing your results
I a saved R file showing results