---
title: "Associations"
author: |
  | Introduction to Quantitative Ecology
  | Fall 2018
  | Chris Sutherland
  | csutherland@umass.edu

header-includes:
   - \usepackage{tabto}
   - \usepackage{verbatim}
   - \usepackage{amsmath}
   - \usepackage{mathtools}
   - \usepackage{graphicx}
   - \usepackage{tikz}
   - \usepackage{tikzpagenodes}
   - \definecolor{OG}{RGB}{0,64,8}
   - \definecolor{LG}{RGB}{0,102,51}
   - \definecolor{myRed}{RGB}{228,26,28}
   - \definecolor{myBlue}{RGB}{55,126,184}
   - \definecolor{myGreen}{RGB}{77,175,74}
   - \definecolor{myPurple}{RGB}{152,78,163}
   - \setbeamercolor{itemize item}{fg=white!0!LG}
   - \setbeamercolor{enumerate item}{fg=white!0!LG}
   - \setbeamercolor{enumerate subitem}{fg=white!70!LG}
   - \setbeamercolor{itemize subitem}{fg=white!70!LG}   
   - \setbeamercolor{itemize subsubitem}{fg=white!70!LG}   
   - \setbeamercolor{navigation symbols}{fg=white!70!LG, bg=white!70!LG}
   - \usepackage{inputenc}
   - \usepackage{booktabs}
   - \usepackage{caption}
   - \usetikzlibrary{patterns,arrows,decorations.pathreplacing}
   - \usepackage{setspace}
   - \DeclareGraphicsExtensions{.pdf,.png,.jpg,.gif}
   - \usepackage{fixmath}
   - \usepackage{tabto}
   - \usepackage{arydshln}
   - \graphicspath{"C:/Users/megha/Box/Courses/NRC_290b_2019_Fall/Images/Assoc/"}

output:
  beamer_presentation:
    highlight: tango
    theme: "default"
    colortheme: "spruce"
    fonttheme: "serif"
    slide_level: 2
    incremental: false
classoption: t
---

```{r setup, echo=FALSE, results='hide', message=FALSE}
library(knitr)
opts_knit$set(root.dir = "C:/Users/megha/Box/Courses/NRC_290b_2019_Fall/")
#opts_knit$set(root.dir = "C:/Users/csutherland/Dropbox/C/Courses class/NRC290b/NRC_290b_2018/")

options(width=120)
options(width.cutoff=80)
opts_chunk$set(prompt = FALSE, comment = NA, mysize=TRUE, size="\\scriptsize")
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knit_hooks$set(document = hook2)
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
```

## Group evaluations

\vfill

1. Specifically for the assignment due tomorrow, how would you describe your comfort levels with R and Excel? 

\begin{enumerate}[\hspace{0.5cm}A.]
  \item[A)] Uncomfortable with both
  \item[B)] Comfortable with Excel, but not R
  \item[C)] Comfortable with R, but not Excel
  \item[D)] Comfortable with both
\end{enumerate}

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2cm,yshift=1cm] at (current page.south east)
{\includegraphics[height=0.25in]{iclicker.png}};
\end{tikzpicture}

\vfill

## One Question

\vfill

1. If you were using Pearson's residuals to see which specific associations in a contingency table were likely to be significant, which of the following values would be the critical value to use?

\begin{enumerate}[\hspace{0.5cm}A.]
  \item[A)] 3.84
  \item[B)] -2 and 2
  \item[C)] 0.05
  \item[D)] 10/25/1980
\end{enumerate}

\vfill

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2cm,yshift=1cm] at (current page.south east)
{\includegraphics[height=0.25in]{iclicker.png}};
\end{tikzpicture}


## Group evaluations

Is there a significant association between salamander \emph{sex} and \emph{habitat type}?

$$\chi^2 = \sum \frac{(\text{Obs} - \text{Exp})^2}{\text{Exp}}$$

1. Calculate $\chi^2$.  
2. Calculate $DF$  
3. Is there a significant association?

\vspace{1cm}

\begin{tabular}{lcccc}
\hline     
           & Dry & Moist & Wet & $\sum$ Row \\ 
\hline
Female     & \emph{370} & \emph{198} & \emph{187} &      \\   
Male       & \emph{359} & \emph{110} & \emph{160} &      \\ 
$\sum$ Col &       &         &       &      \\ 
\hline
\end{tabular}

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2cm,yshift=-1.5cm] at (current page.east)
{\includegraphics[height=1.75in]{chiTab.jpg}};
\end{tikzpicture}

## Associations

What are associations?

- dealing with two categorical variables
- known as a *contingency table*
    - data are *cross tabulated* frequencies
    - each cell represents a count

## Associations

- dealing with two categorical variables
- known as a *contingency table*
    - data are *cross tabulated* frequencies
    - each cell represents a count

\vspace{0.5cm}

\centering

\begin{tabular}{lccc}
\hline     
Salamander & Upland & Wetland & Floodplain \\ \hline
Present    & \emph{38}    & \emph{30} & \emph{24} \\   
Absent   & \emph{12}    & \emph{20} & \emph{26} \\ 
\hline
\end{tabular}

\vspace{1cm}

## Associations

- dealing with two categorical variables
- known as a *contingency table*
    - data are *cross tabulated* frequencies
    - each cell represents a count
- visualize using a *bar chart* 

```{r, echo=FALSE, message=FALSE, results="hide"}
mm <- matrix(c(38,30,24,12,20,26),2,3,byrow=T)
pdf("Images//Assoc//manderContTab.pdf",height=4,width=6)
par(oma=c(0,0,0,0),mar=c(4,4,2,2))
barplot(mm, names=c("Upland","Wetland","Floodplain"),
        beside=T, col=adjustcolor(c(2,3),0.2), las=1,ylab="Count", ylim=c(0,45))
legend("topright",c("Present", "absent"),pch=22,bg=adjustcolor(c(2,3),0.2),bty="n")
dev.off()
```

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=2.5cm] at (current page.south)
{\includegraphics[height=2in]{manderContTab.pdf}};
\end{tikzpicture}

## Associations

The key question with contingency tables:

- is there a significant association between the categorical variable A and categorical variable B?
- are these numbers different to what we would expect to see by chance?

## Associations

The key question with contingency tables:

- is there a significant association between the categorical variable A and categorical variable B?
- are these number different to what we would expect to see by chance?
- answer using the *Chi-squared test*

$$ \chi^2 = \sum \frac{(O-E)^2}{E}$$

- O: the observed data
- E: the expected value if there was *no association*
- $\chi^2$: the test statistic

## Associations

In a Chi-squared test from the following contingency table, I get a test statistic of $\chi^2 =  8.32$. Is there a significant association between spotted salamander presence and vernal pool type at the 5% level?

\vspace{0.5cm}

\centering

\begin{tabular}{lccc}
\hline     
Salamander & Upland & Wetland & Floodplain \\ \hline
Present    & \emph{38}    & \emph{30} & \emph{24} \\   
Absent   & \emph{12}    & \emph{20} & \emph{26} \\ 
\hline
\end{tabular}

\vspace{1cm}

\begin{enumerate}[\hspace{0.5cm}A.]
  \item[A)] Yes, its significant!
  \item[B)] No, it's not significant!
\end{enumerate}

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2cm,yshift=2cm] at (current page.south east)
{\includegraphics[height=1.4in]{chiTab.jpg}};
\end{tikzpicture}

## Associations

In a Chi-squared test from the following contingency table, I get a test statistic of $\chi^2 =  8.32$. Is there a significant association between spotted salamander presence and vernal pool type at the 5% level?

\vspace{0.5cm}

\centering

\begin{tabular}{lccc}
\hline     
Salamander & Upland & Wetland & Floodplain \\ \hline
Present    & \emph{38}    & \emph{30} & \emph{24} \\   
Absent   & \emph{12}    & \emph{20} & \emph{26} \\ 
\hline
\end{tabular}

\vspace{1cm}

\begin{enumerate}[\hspace{0.5cm}A.]
  \item[A)] \textbf{Yes, its significant!}
  \item[B)] \textcolor{gray}{No, it's not significant!}
\end{enumerate}

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2cm,yshift=2cm] at (current page.south east)
{\includegraphics[height=1.4in]{chiTab.jpg}};
\end{tikzpicture}

## Associations - observed

$$ \chi^2 = \sum \frac{(O-E)^2}{E}$$

The observed values ($O$):

- the data we observe (obviously!)
- cross tabulated counts

## Associations - expected

$$ \chi^2 = \sum \frac{(O-E)^2}{E}$$

The expected values ($E$):

- the data we would expect by change
- the data we would expect if there was no association

$$E = \frac{\text{row total} \cdot \text{col total}}{\text{grand total}}$$

## Associations - expected

$$ \chi^2 = \sum \frac{(O-E)^2}{E}$$

Degrees of freedom:

$$ DF = (no. columns -1) \times (no. rows -1)$$

## Associations - expected

$$ \chi^2 = \sum \frac{(O-E)^2}{E}$$

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=3.5cm] at (current page.south)
{\includegraphics[height=2in]{chi5.png}};
\end{tikzpicture}

## Associations - invertebrate example

Invertebrate group habitat selection:

\vfill

\centering

\begin{tabular}{lccc|c}
\toprule     
           & Ant & Bug  & Beetle & Total \\ \hline
Upper leaf & 15  &  13  & 68     &  96   \\   
Lower leaf & 12  &  11  & 15     &  38   \\   
Stem       & 65  &  78  &  5     & 148   \\   
Bud        &  3  &  21  &  3     &  27   \\ \hline    
Total      & 95  & 123  & 91     & 309   \\ \bottomrule    

\hline
\end{tabular}

\vfill

## Associations - invertebrate example

First we need to state the hypotheses!

- Null hypothesis:

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=5cm] at (current page.south)
{\includegraphics[height=1.5in]{QM.png}};
\end{tikzpicture}

## Associations - invertebrate example

First we need to state the hypotheses!

- Null hypothesis:
- the *no habitat preference* hypothesis
- random!

"There is no association between invertebrate group and habitat"

## Associations - invertebrate example

First we need to state the hypotheses!

- Null hypothesis:
- the *no habitat preference* hypothesis
- random!

"There is no association between invertebrate group and habitat"

\vspace{0.5cm}

- Alternative hypothesis:

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=2cm] at (current page.south)
{\includegraphics[height=1.5in]{QM.png}};
\end{tikzpicture}

## Associations - invertebrate example

First we need to state the hypotheses!

- Null hypothesis:
- the *no habitat preference* hypothesis
- random!

"There is no association between invertebrate group and habitat"

\vspace{0.25cm}

- Alternative hypothesis:
- the *habitat preference* hypothesis
    - direction not explicitly stated
    - can be positive or negative
- not random!

"There *is* an association between invertebrate group and habitat"

## Associations - invertebrate example

\vfill
\centering
\Large

Demo in \texttt{Excel}

\vfill

## Associations - invertebrate example

- The test statistic is $\chi^2 = 146.98$. 
- What do we conclude?

## Associations - invertebrate example

- The test statistic is $\chi^2 = 146.98$. 
- What do we conclude?
    - $p < 0.05$
    - reject the null hypothesis
    - accept the alternative hypothesis
    - there *is* a significant association between inverts and habitat!
    
## Associations - invertebrate example

- The test statistic is $\chi^2 = 146.98$. 
- What do we conclude?
    - $p < 0.05$
    - reject the null hypothesis
    - accept the alternative hypothesis
    - there *is* a significant association between inverts and habitat!
- BUT! which associations are significant? 

\centering

\begin{tabular}{lcccc}
\toprule     
           & Ant & Bug  & Beetle & Total \\ \hline
Upper leaf & 15  &  13  & 68     &  96   \\   
Lower leaf & 12  &  11  & 15     &  38   \\   
Stem       & 65  &  78  &  5     & 148   \\   
Bud        &  3  &  21  &  3     &  27   \\ \hline    
Total      & 95  & 123  & 91     & 309   \\ \bottomrule    

\hline
\end{tabular}

## Associations - significant associations

Two ways we can evaluate which associations are likely to be significant:

1. Cell-specific $\chi^2$ values
    - greater than 3.8 is likely to be significant
    - 3.8 is significant test statistic with 1 degree of freedom

## Associations - significant associations

Two ways we can evaluate which associations are likely to be significant:

1. Cell-specific $\chi^2$ values
    - greater than 3.8 is likely to be significant
    - 3.8 is significant test statistic with 1 degree of freedom
    
2. Pearson residuals
    - provides sign of association
    - provides relative size of the association
    - if residual is >2 or < -2 then likely to be significant 

$$ Residual = \frac{\text{Observed}-\text{Expected}}{\sqrt{\text{Expected}}}$$

## Associations - invertebrate example

\vfill
\centering
\Large

Back to \texttt{Excel}

\vfill


## Analyzing contingency table data in \texttt{Excel}

Two `Excel` functions for doing Chi-squared or Goodness of fit tests but no method in *Analysis Tool Pack*:

## Analyzing contingency table data in \texttt{Excel}

Two `Excel` functions for doing Chi-squared or Goodness of fit tests but no method in *Analysis Tool Pack*:

- `CHITEST(`*observed*,*expected*`)`
    - must calculate the *expected* values
    - must be in the same table format

## Analyzing contingency table data in \texttt{Excel}

Two `Excel` functions for doing Chi-squared or Goodness of fit tests but no method in *Analysis Tool Pack*:

- `CHITEST(`*observed*,*expected*`)`
    - must calculate the *expected* values
    - must be in the same table format

- `CHIDIST(`*Chi value*,*derees of freedom*`)`
    - must calculate $\chi^2$ and DF
    - just calculates a *p*-value

\vfill
\centering

Demo in \texttt{Excel}

\vfill

## Analyzing contingency table data in \texttt{R}

- Chi-square test in `R`
    - `chisq.test(`*contingency table*`)`
    - data must be formatted like a contingency table
    - can make a `data.frame` 
    
```{r}
# a dataframe
tab.df <- data.frame(Ant = c(15,12,65,3),
                     Bug = c(13,11,78,21),
                     Beetle = c(68,15,5,3))
rownames(tab.df) <- c("Upper", "Lower", "Stem", "Bud")
tab.df
```

## Analyzing contingency table data in \texttt{R}

- Chi-square test in `R`
    - `chisq.test(`*contingency table*`)`
    - data must be formatted like a contingency table
    - can make a `matrix` 
    
```{r}
# a dataframe
tab.mat <- matrix(c(15,12,65,3,
                    13,11,78,21,
                    68,15,5,3), nrow=4, ncol=3, byrow=FALSE)
rownames(tab.mat) <- c("Upper", "Lower", "Stem", "Bud")
colnames(tab.mat) <- c("Ant", "Bug", "Beetle")
tab.mat
```

## Analyzing contingency table data in \texttt{Excel}

- Chi-square test in `R`
    - `chisq.test(`*contingency table*`)`
    - data must be formatted like a contingency table
    - can make a `data.frame` or a `matrix` 
    
```{r}
# conduct the Chi-square test
chisq.test(tab.df)
```

## Analyzing contingency table data in \texttt{Excel}

- Chi-square test in `R`
    - `chisq.test(`*contingency table*`)`
    - data must be formatted like a contingency table
    - can make a `data.frame` or a `matrix` 
    
```{r}
# extract the expected values
chisq.test(tab.df)$expected
```

## Analyzing contingency table data in \texttt{Excel}

- Chi-square test in `R`
    - `chisq.test(`*contingency table*`)`
    - data must be formatted like a contingency table
    - can make a `data.frame` or a `matrix` 
    
```{r}
# calculate the Pearson residuals
obs <- tab.df
exp <- chisq.test(tab.df)$expected
(obs-exp) / sqrt(exp)
```

## Analyzing contingency table data in \texttt{Excel}

What can we conclude from our analysis of the invertebrate data using the Chi-square test for association?     
```{r}
# conduct the Chi-square test
chisq.test(tab.df)

# Pearson residuals
(obs-exp) / sqrt(exp)
```


## 'Pioneer Valley Camera Trapping Data'

The data:

- 142 camera traps placed throughout the valley
- each camera has 2 categorical covariates:
    - land use: 'altered' (A), 'natural' (N) and 'urban' (U)
    - scent lure: 'Badlands Bob' (BB) and 'Powder River' (PR)
- we will focus on four species:
    - bobcat *lynx rufus*
    - domestic cat *felis catus*
    - coyote *canis latrans*
    - domestic dog *canis familiaris*

## Group exercise

Using the `Pioneer Valley Camera Trapping Data' investigate whether there is a statistically significant association between the four species and:

1. habitat type
2. scent lure

(i.e., conduct two analyses)

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=0cm,yshift=2.5cm] at (current page.south)
{\includegraphics[height=0.8in]{camtraps.png}};
\end{tikzpicture}
    
<!---
## Group exercise

Write a mini-report that includes the following for both tests:

- the biological question
- a table and a graph of the data
- the null and alternative hypothesis
- details of the statistical test:
    - which test
    - test statistic
    - degrees of freedom
    - *p*-value
- whether there a statistical association and if so which were likely to be significant
- a biological interpretation of the results

Submit as a **PDF** by midnight Wednesday 16$^{th}$ November


## Question time!

\vfill

1. If you were using Pearson's residuals to see which specific associations in a contingency table were likely to be significant, which of the following values would be the critical value to use?

\begin{enumerate}[\hspace{0.5cm}A.]
  \item[A)] 3.84
  \item[B)] -2 and 2
  \item[C)] 0.05
  \item[D)] 10/25/1980
\end{enumerate}

\vfill

\begin{tikzpicture}[remember picture,overlay]
\node[xshift=-2cm,yshift=1cm] at (current page.south east)
{\includegraphics[height=0.25in]{iclicker.png}};
\end{tikzpicture}

--->

